An atypical basement membrane forms a midline barrier in left-right asymmetric gut development

Version 1

Abstract

Abstract

Correct intestinal morphogenesis depends on the early embryonic process of gut rotation, an evolutionarily conserved program in which a straight gut tube elongates and forms into its first loops. However, the gut tube requires guidance to loop in a reproducible manner. The dorsal mesentery (DM) connects the gut tube to the body and directs the lengthening gut into stereotypical loops via left-right (LR) asymmetric cellular and extracellular behavior. The LR asymmetry of the DM also governs blood and lymphatic vessel formation for the digestive tract, which is essential for prenatal organ development and postnatal vital functions including nutrient absorption. Although the genetic LR asymmetry of the DM has been extensively studied, a divider between the left and right DM has yet to be identified. Setting up LR asymmetry for the entire body requires a Lefty1+ midline barrier to separate the two sides of the embryo—without it, embryos have lethal or congenital LR patterning defects. Individual organs including the brain, heart, and gut also have LR asymmetry, and while the consequences of left and right signals mixing are severe or even lethal, organ-specific mechanisms for separating these signals are not well understood. Here, we uncover a midline structure composed of a transient double basement membrane, which separates the left and right halves of the embryonic chick DM during the establishment of intestinal and vascular asymmetries. Unlike other basement membranes of the DM, the midline is resistant to disruption by intercalation of Netrin4 (Ntn4). We propose that this atypical midline forms the boundary between left and right sides and functions as a barrier necessary to establish and protect organ asymmetry.

Introduction

Humans and most other animals are bilaterians—animals whose left and right external features can be mirrored—but often the internal organs have striking LR asymmetries. For example, in humans the heart resides on the left side of the thoracic cavity, the liver is predominantly on the right, and the spleen is on the left. Even paired organs like the lungs can show LR asymmetries— the left human lung has two lobes while the right lung has three. The left and right sides of the body are specified early in development, after the anterior/posterior and dorsal/ventral axes have been established.1,2 This patterning relies heavily on the expression of Sonic hedgehog (Shh) upstream of Nodal on the left side, and the repression of these genes by Activin on the right side.3 The localization of SHH-producing cells to the left is accomplished by nodal flow (mouse,4,5 zebrafish6 and Xenopus7 embryos) or rotational cell movements around the node (chicken and pig embryos),8 as well as cell death at the embryonic midline that may be a consequence of its abundant extracellular matrix (ECM).9

Establishing the vertebrate LR body axis depends on a midline barrier to separate side-specific diffusible signals.10–15 This is achieved with a highly specific expression pattern of an inhibitor, Lefty1, at the center of the embryo which prevents the diffusible, left-sided signal NODAL from crossing to the embryo’s right side.11–15 Sixty percent of Lefty1-knockout mouse embryos die in utero and an additional 20% die before weaning, suffering from left isomerism of the lungs (in other words, both lungs have left lung lobation) and positional defects of the heart and the major vessels leading into/out of it.13 Other important examples of laterality defects are seen in conjoined twins, in which an embryo divides partially at the primitive streak stage (Day 13-14 of gestation for humans).16 In laterally conjoined (dicephalus) twins which form when the two primitive streaks are parallel, the left side of one embryo and the right side of the other are connected without the LEFTY1 barrier in between. Consequently, these conjoined twins often exhibit LR defects.17,18 This underscores that separation of left and right signals is fundamental in early development for setting up correct placement and LR asymmetric patterning of individual organs including the heart and gut.6,10,19–21 Of all the organs with LR asymmetry, only the brain is known to harbor an organ-specific midline barrier.22–28 Thus, midline barriers may have broad developmental significance for the embryo and its organs, yet very few such structures have been characterized.

Even organs that do not have obvious LR differences in the adult develop as a result of conserved LR asymmetric morphogenesis. The intestine is an excellent model for this, especially given the relatively simple tubular structure of the organ itself. During development, the intestine grows to great lengths (about eight meters in adult humans),29 and this long tube must be looped to fit into the body cavity in a stereotypical, species-specific way.30,31 When the developmental program directing gut looping is perturbed, as is the case for one in 500 infants who have congenital malrotation of the gut, there is an increased risk of volvulus, a lethal self-strangulation of the gut that requires immediate pediatric surgical intervention.32

The gut tube does not loop autonomously. Rather, gut looping is directed by the neighboring dorsal mesentery (DM) (Fig. 1A), a thin mesodermal organ that connects the gut to the rest of the body and through which intestinal blood and lymphatic vessels traverse.31,33–35 The left and right sides of the DM take on different properties at the molecular, cellular, and extracellular levels which are critical to initiate asymmetric gut looping and vascular morphogenesis.33,36–40 Gut tilting is the symmetry-breaking event that initiates asymmetric gut looping, which occurs at embryonic day 10.5 (E10.5) in the mouse and Hamburger-Hamilton stage 19-21 (HH19-21)41 in the chicken embryo (Fig. 1B).37 Gut tilting is driven by the condensation of the ECM in the left DM, and an expansion of the ECM on the right (Fig. 1B).36 Blood vasculature also develops asymmetrically. Although endothelial precursors exist on both the left and right DM prior to tilting, as the asymmetries are established the right-sided endothelial cells emigrate rather than forming vessels.33 Only a left-sided gut artery develops further (Fig. 1B), which goes on to supply blood to a significant portion of the adult intestine (Fig. 1A).33 Interestingly, these right-sided endothelial cells emigrate dorsally and ventrally, but they cannot cross directly over to the left side.33 This might indicate the presence of a barrier against cell migration at the midline.

LR asymmetry in the DM is critical for proper gut looping and vascular patterning.

(A) Asymmetries in the DM drive the formation of the first and subsequent gut loops. Concurrently, the vasculature is being patterned in the DM. The 1° longitudinal artery (*) gives rise to the ileocolic artery, which provides a significant portion of the adult intestine with critical blood flow. (B) HH17: The DM has cellular symmetry. HH18 and 19: The right mesenchyme begins expanding and the right epithelial cells elongate. The right-sided endothelial cells (vascular precursors) begin to disperse and leave the compartment, while the left-sided endothelium is maintained to become the future gut arteries. HH22: The asymmetric forces have pushed the gut tube to the left. A left-sided blood vessel, the 1° longitudinal artery (*) has formed. GT = gut tube, DM = dorsal mesentery, DA = dorsal aorta, SMA = superior mesenteric artery.

The differences in the left and right sides of the DM are well understood at multiple levels of biology, but the maintenance of these asymmetries has not been explored. The classes of genes with asymmetric expression in the DM include transcription factors, ECM components, and genes which encode diffusible signals that could spread within the DM compartments to create morphogen gradients across the LR axis. Instead of a LR gradient, however, there is a sharp delineation between left and right DM in gene expression and protein localization, as well as LR asymmetric distribution of extracellular glycosaminoglycans, such as hyaluronan.33,36–40 Not surprisingly, perturbing even just one gene’s asymmetric expression patterns in the DM leads to aberrant gut looping patterns and abnormal vascular lesions.33,38–40 Thus, we hypothesized that a barrier exists in the center of the DM to segregate cells and diffusible signals to the left and right compartments. Here, we show that an atypical basement membrane forms a boundary between left and right cell populations and limits diffusion across the DM midline.

Results

Cells of left- and right-origin meet but do not mix in the DM

The DM mesenchyme forms by bilateral epithelial-to-mesenchymal transition (EMT) and ingression of coelomic epithelium, which is derived from the splanchnic mesoderm; the left DM comes from the left coelom and the right DM arises from the right coelom (Fig. 2A and B).35–37 This was visualized by injecting early chicken embryos with DiI and DiO into the right and left coelomic cavities, respectively (Fig. 2C-F), or by electroporating each side with plasmids encoding different fluorophores (Fig. 2G and H). While the left and right cells meet at the middle of the DM, they never cross over or mix.33,36–38,40,42–45 This striking separation occurs despite the lack of a visible histological boundary between the two sides as shown by H&E staining at HH20 and HH21, when the left is condensing and the right is expanding to drive the leftward gut tilting (Fig. 2K and L). Early in development the endoderm effectively separates the left and right splanchnic mesoderm (Fig. 2I and J), but once the DM forms and the endoderm descends it is likely important to continue separating the two sides until asymmetries can be established (Fig. 1B). In support of this, we have previously shown that when cell-cell adhesion is interrupted in the left DM, the cells become more dispersed36,38 and extend filopodia over towards the right side, suggesting pathogenic cell migration.38 Thus, the critical separation between left and right cells in the DM can be disrupted, necessitating a mechanism for protecting these asymmetries.

During DM formation from EMT and ingression of the coelomic epithelia, right and left cells do not mix.

When the coelomic cavity is injected with DiI at HH12-13 (A), the labeled cells give rise to the mesenchymal and epithelial cells of the DM on the corresponding side of the embryo (B). When DiI and DiO are injected at HH12-13 into left and right coeloms, respectively (C), labeled cells are still segregated at HH18 (D, E and F). The same results are found when cells are labeled by electroporation with pCAG-GFP (left) and pCl-H2B-RFP (right) (G and H), both when the midline is continuous (HH19, G) and once it has disappeared (HH21, H). (I-L) H&E staining of the DM at HH18 (I) shows “empty space” between the notochord, endoderm, and dorsal aortae. At HH19 (J), this space gains some cells (arrows), and the space is completely filled in by HH20 (K) and HH21 (L). Scale bars = 60 um. nt = neural tube, c = coelom, ao = aorta, N = notochord, s = somite, DM = dorsal mesentery, L = left, R = right.

The DM midline is not marked by Lefty1, but by laminin

The early embryo uses a molecular barrier of Lefty1-expressing cells to separate laterality signals so the LR axis is established correctly.11–15 To see if this mechanism is adapted by the intestine for establishment of its laterality, we performed Lefty1 RNA in situ hybridization on both early (HH9) and later (HH19) stages. While Lefty1 was expressed at the midline of early embryos as expected (Fig. 3A), it was not expressed at the midline of the DM at HH19 (Fig. 3B). This indicates that a different mechanism must be at work during the establishment of gut asymmetries. Interestingly, scanning electron microscopy (SEM) data showed a fibrous matrix between the notochord and endoderm where the DM will later develop, suggesting that ECM may separate the two sides before they coalesce into the DM (Fig. 3C and D). Consistent with this hypothesis, basement membranes are found in other biological contexts where a barrier is needed, such as in the skin or around blood vessels.46 Basement membranes are dense ECM requiring laminin, collagen IV, nidogen, and perlecan and/or agrin (both heparan sulfate proteoglycans) with a large variety of other components that can be integrated to create specific “flavors” of basement membrane tuned to different barrier contexts.46 We therefore postulated that the DM midline has a physical barrier consisting of basement membrane, rather than a Lefty1 molecular barrier.

The DM midline is marked by laminin immunofluorescence.

(A) Lefty1 expression is seen at the embryonic midline of this HH9 embryo (arrow). (B) Lefty1 expression is not seen at the midline of the DM (dashed box) of an HH19 embryo. (Notochord=positive control.) (C, D) Scanning electron microscopy (SEM) images of a fixed embryo at HH15-16 show fibrous ECM between the notochord and endoderm. C is from a more posterior axial level than D. (E-H) Time course of midline dynamics from HH18-HH21, marked by laminin. Scale bars = 25 µm. (I) Quantification of the intensity of laminin immunofluorescence over development, normalized to laminin staining around the neural tube. Biological replicates: HH18 n = 2, HH19 n = 3, HH20 n = 3, and HH21 n = 2. Each dot represents one image quantified. Statistical analyses are unpaired Welch’s t-tests. HH18-19: p = 0.0188, HH19-20: p = 0.0118, HH20-21: p = 0.0003. Additionally, there is a significant (p < 0.0001) linear trend among the means from HH18 to HH21 with a slope of −0.1019, r2 = 0.5906. (J) Midline appearance from hindgut to foregut in an HH19 embryo, marked by laminin. Scale bars = 25 µm. (K) Quantification of DM midline length of three HH19 embryos, from the separation of the notochord and endoderm (hindgut) to the fusion of the aortae (foregut, coinciding with midline fragmentation). (L) Model of DM midline time kinetics. N = notochord, E/en = endoderm, Ao = aorta, nt = neural tube, c = coelomic cavity, CE = coelomic epithelium, L = left, R = right.

To test this hypothesis, we visualized the basement membrane marker laminin by immunofluorescence (IF) for laminin alpha 1 (Lama1) at developmental stages where DM asymmetries are being established. As expected, this marked several typical, single-layer basement membranes underlying polarized cells, such as around the notochord (Fig. 3E and F),47,48 coelomic epithelium (Fig. 3E-H),49 and gut endoderm (Fig. 3E-H).50 We also observed scattered laminin staining in the DM mesenchyme, which is a consequence of those cells carrying basement membrane fragments with them after EMT and ingression from the coelomic epithelium.51,52 Interestingly, we identified a previously uncharacterized atypical double basement membrane within the DM. At HH18, when cellular DM asymmetries are first being initiated at the level of the midgut (which forms the small intestine), laminin IF marked an oval-shaped structure just ventral to the notochord and dorsal to the gut endoderm (Fig. 3E). No cells were seen within this structure as shown by a lack of nuclear staining (Fig. 3E) and empty space in SEM (Fig. 3C and D, S1C). As the DM elongates and asymmetries become more apparent (HH19), this midline structure lengthened, forming two parallel lines connecting the endoderm and notochord (Fig. 3F). One stage later (HH20), the midline was still present but began to appear fragmented (Fig. 3G). By HH21, the asymmetries of the DM are established—the right DM has expanded, the left DM has condensed, the gut has tilted to the left, and vascular precursor cells have been driven out of the right side (Fig. 1B).33,40,43 Surprisingly, the midline disappeared by this stage, while laminin IF underlying the coelomic epithelium and endoderm remained intense (Fig. 3H). The lengthening of the midline and its subsequent loss occurred in an anterior-to-posterior wave down the embryonic gut tube (Fig. 3J and K). Consequently, anterior sections of younger embryos (i.e., HH12-13) had similar midline structures to posterior, older sections (i.e., HH18-19) (compare Fig. S2E-H with S2A-D).

The DM midline consists of a transient, true basement membrane

The combination of laminin, nidogen, perlecan (or agrin), and collagen are the foundation of all basement membranes.46 To further characterize the nature of the ECM at the DM midline, we did IF staining for nidogen and perlecan, confirming co-localization with laminin at the midline barrier (Fig. S3A, S3B). This further illustrates that the DM midline consists of basement membrane.

In addition to the four foundational basement membrane components, a myriad of other proteins, proteoglycans, and glycoproteins can assemble onto the basement membrane.53 Consequently, there is a vast variety of “flavors” of basement membrane with different physical properties and different signals to adjacent cells about polarity, migration, or other behaviors.53 A common basement membrane constituent is fibronectin, which is best known for its role in the provisional matrix during wound healing.54 At our developmental stages of interest, fibronectin localized to the dorsal aorta which often coincides with the most ventral part of the midline (Fig. S2, S3C). Together, we model the midline as a transient double basement membrane that bisects the DM during developmental stages when critical asymmetries are being established (Fig. 3L).

The midline does not originate from the left or right DM

Although we have established the time kinetics of DM midline formation, the origin of this structure remains elusive. The midline is sandwiched between mesenchymal cells from the left and right DM, unbiased to either the left or the right side (Fig. 2F and G). Intriguingly, mesenchymal cells like these are not usually competent to construct an organized basement membrane.46,55 Mesenchymal cells can secrete matrix components,50 but the organization of these components into a basement membrane is dependent on the presence of cell surface anchors which are characteristic of tissues like polarized epithelium or endothelium, not mesenchymal cells.55 In the case of LAMA1, it is known that this protein is secreted by the epithelia in the developing intestine, not the mesenchyme.50 Indeed, RNA in situ hybridization for Lama1 did not show enriched expression in the mesenchyme at the DM midline (Fig. 4B and C). Moreover, if the cells adjacent to the midline were secreting and organizing the basement membrane, we would expect these cells to be polarized like the cells of the coelomic epithelium or endoderm. As expected, the left coelomic epithelium was polarized at HH19 relative to its basement membrane as quantified by Golgi staining with GM130 (Fig. 4D and E).38 However, GM130 staining and quantification showed that cells immediately to the left or right of the midline have random orientation (Fig. 4D and E). Together, these data allow us to rule out a mesenchymal origin for the DM midline.

The midline basement membrane is not made by the DM mesenchyme or EMT of the endoderm, and the notochord is not suffcient for midline formation.

(A-C) Lama1 RNA in situ hybridization and adjacent sections with laminin IF staining (A’-C’). Scale bars = 50 um. (D, E) Cell polarity analysis from GM130 staining shows that the mesenchymal cells immediately to the left or right of the midline (“left/right margin”) and within the double membrane (“midline cells” have random polarity, as do the cells of the right mesenchyme (random polarization control); in contrast to the strong apical-basal polarity in cells of the left coelomic epithelium. Five embryos were used for these quantifications. Number of cells per circle histogram: left coelomic epithelium = 209, left margin = 346, midline cells = 118, right margin = 413, right mesenchyme = 514. (F) Electroporation mix containing pCAGEN-GFP plasmid was injected under an HH14/15 embryo and then electroporated to specifically target the endoderm. Lineage tracing endoderm-derived cells to HH19 (G) and HH21 (H) by pCAGEN-GFP electroporation of the endoderm. (I) Model of notochord transplant method. A piece of notochord (red) was isolated from an HH14 donor embryo. In a stage-matched recipient, a cut was made adjacent to the neural tube and the donor notochord was inserted into this slit. (J, L) RNA in situ hybridization for Chordin to mark the native notochord (blue dashed circle) and transplanted notochord (red dashed circle). (K, M) Laminin immunohistochemistry to mark basement membrane including the midline (white arrow). Notochords are marked with an N (native notochord) and N* (transplanted notochord). (J) and (K) are from the same embryo, as are (L) and (M). n = 8, scale bars = 50 um. GT = gut tube, DM = dorsal mesentery, E = endoderm.

During endodermal descent, endodermal cells are not left behind to form the midline

Given that the DM midline connects the notochord and endoderm, both of which have their own basement membranes and are very closely associated early in development, we hypothesized that one or both of these structures contributes to midline production (Fig. 3C/E/J, S1A). As the embryo grows and the DM elongates, the distance between the notochord and endoderm increases (Fig. 2I-J, 3E-H), and the midline is found between them as a double line of basement membrane. Thus, we hypothesized that as the endoderm descends ventrally, it undergoes EMT and leaves behind basement membrane-carrying cells to form the midline. To test this, we developed a method to specifically target the endoderm using DNA electroporation (Fig. 4F). Briefly, we injected pCAG-GFP plasmid underneath HH14-15 embryos and applied an electric pulse such that the endodermal cells would take up the DNA (Fig. 4F), so we could lineage trace endodermal cells during DM formation. Interestingly, embryos isolated at HH19 and HH21 showed that GFP-labeled cells remain restricted to the endoderm—there were no GFP+ mesenchymal cells present in the DM (Fig. 4G, H). This indicates that the DM midline is not formed from EMT of basement membrane-carrying endodermal cells.

The notochord is not sufficient for DM midline formation

To test whether the notochord is sufficient for midline formation, we performed notochord transplant experiments. In brief, the notochord was removed from an HH12-15 embryo. An incision was made in a stage-matched recipient embryo adjacent to the neural tube and the donor notochord was inserted into this slit (Fig. 4I). Embryos continued to develop until isolation at HH19. These transplants were done to the left and right sides of different embryos (Fig. 4J/K and 4L/M respectively). RNA in situ hybridization was performed for chordin to ensure that the transplanted notochord was alive and functioning (Fig. 4J and L).56,57 Laminin IF did not reveal a secondary midline-like structure associated with the ectopic notochord (Fig. 4K and M) while the normal midline was unaffected (Fig. 4K and M, white arrows). This result is seen regardless of whether the transplants are done to the embryo’s left or right side. From this, we conclude that the notochord is not sufficient for the formation of DM midline.

The DM midline is resistant to degradation by Netrin4

Laminin matrices are susceptible to competitive disruption by the matrix protein Netrin4 (NTN4).58,59 NTN4 has very high binding affinity for laminin gamma subunits, such that NTN4 can prevent the formation of new laminin networks, which are the foundation upon which other basement membrane components assemble, and can also disrupt existing laminin networks.58,59 Ntn4 is not endogenously expressed in the DM (data not shown), which allows us to use it as a tool to target basement membranes in the DM. As expected, when we overexpressed Ntn4 on either side of the DM by electroporation, we perturbed the basement membrane underlying the coelomic epithelium and depleted the scattered laminin staining in the mesenchyme that results from EMT creating the DM (Fig. 5B, D, and E vs. controls 5A and C).51,52 Intriguingly, the DM midline basement membrane remained visibly intact in all embryos with lateral (Fig. 5B and D) and bilateral (Fig. 5E) Ntn4 overexpression in the DM. Similarly, endodermal Ntn4 overexpression caused much less disruption of the endodermal or midline basement membranes (Fig. 5G vs control 5F) when compared to its effect on coelomic epithelium basement membrane. This was true even when the electroporations were done much earlier in development (HH10 and HH12-13, data not shown). This suggests that the midline and endoderm may have basement membranes of the same “flavor,” possibly pointing to a common origin. In contrast, the basement membrane beneath the coelomic epithelium may be more susceptible to NTN4 disruption because of the prior EMT-induced breaks in the basement membrane,51,52 or a different protein composition.

Ectopic expression of Netrin4 by electroporation visibly affects the basement membrane underlying the coelomic epithelium, but not that underlying the endoderm, or the midline.

(A, C) Electroporation of the left (A) or right (C) DM with the control, pCAG-GFP, had no effect on the basement membrane of the coelomic epithelium (arrows). Left n = 6. Right n = 4. (B, D) Electroporation of mouse Netrin4 (pCAGEN-Ntn4) and pCAG-GFP into the left (B) or right (D) DM disrupted the basement membrane underlying the coelomic epithelium (arrows) and scattered basement membrane in the mesenchyme (asterisks). Left n = 5. Right n = 5. (E) Double DM electroporations also disrupt the coelomic epithelium (arrows) but the midline appears intact despite being contacted by Ntn4+ cells (n = 3). (F, G) Electroporation of the endoderm directly with pCAG-GFP or pCAGEN-Ntn4 does not visibly affect the basement membrane underlying the endoderm (open round pointers). Control n = 5. Ntn4 n = 8. The midline appears unaffected by any of these perturbations. Scale bars = 100 um.

DM midline is a barrier against diffusion

Genes including Cxcl1233,39 and Bmp440 which encode diffusible signals are expressed asymmetrically in the DM (Fig. 6A and B). So too are genes encoding enzymes that are secreted into the ECM, like the HA-modifying enzyme TSG6.39 The expression domains of these genes have a sharp boundary at the midline, since left and right cells do not mix. However, the secreted protein products of these genes may be able to diffuse across the DM if their movement is not limited (Fig. 6C). We know that experimentally mixing left and right signals is detrimental to gut tilting and vascular patterning33,36–40—for example, ectopic expression of pro-angiogenic Cxcl12 on the right side results in an aberrant vessel forming on the right.33 Moreover, when the CXCR4 receptor antagonist AMD3100 (MW=502.78) is introduced to the left DM, it abolishes vascular development on the left. However, when the same drug is introduced to the right DM, the left-sided vascular development remains intact.33 This phenomenon suggests a barrier against diffusion.

The DM midline serves as a barrier against diffusion.

(A, B) Genes encoding diffusible signals including Cxcl12 and Bmp4 are expressed asymmetrically in the DM. (C) Hypothesis for the role of the midline in limiting diffusion of left and right signals across DM. (D) At HH19, the midline is intact (white arrow) and diffusion of 3000 MW dextran (green) is limited to the right side. n = 4/4. (E) At HH20, the midline (white arrow) has begun to fragment. Diffusion across the midline is prohibited in some embryos (n = 2/9) but permitted in others (n = 7/9). (F) At later stages when the midline has disappeared, diffusion is allowed through the DM (n = 7/7). (G) Schematic of dextran injections into the right DM. (H) A dextran injection being performed on an HH23 embryo in ovo. (I) Proportion of dextran-labeled cells in the left vs. right DM, with unpaired t test. Scale bars = 50 um. LDM = left dorsal mesentery. RDM = right dorsal mesentery. GT = gut tube. E = endoderm. L = left. R = right. N = notochord. DA = dorsal aorta.

To test if the basement membrane structure at the midline is forming a functional barrier against diffusion, we injected 3 kDa fluorescent dextran directly into the right side of the DM (Fig. 6G and H). When these injections are performed at stages where the midline is intact (HH19), movement of dextran through the tissue was limited to the right side (n = 4/4 embryos) (Fig. 6D). When the basement membrane midline appears fragmented (HH20), these injections produced mixed results—in some embryos (n = 2/9), diffusion across the midline was prevented and in others (n = 7/9) the dextran was able to move into the left mesenchyme of the DM (Fig. 6E). Finally, at stages where no organized basement membrane structure remains at the midline (HH23), diffusion of dextran was always permitted across the entire width of the DM (n = 7/7) (Fig. 6F). Collectively, these data suggest that the basement membrane structure at the midline forms a transient functional barrier against diffusion (Fig. 6I) as well as a boundary for the left and right cells.

Discussion

Establishing the left and right body plan in the early embryo is a fundamental part of development and this process depends on the presence of a Lefty1+ midline barrier. Individual organs, too, have LR asymmetries, but the brain has the only known organ-specific midline barrier, where commissural axons of the brain and spinal cord are tightly controlled by midline-localized guidance and repulsion cues including FGFs, SLIT/ROBO signaling, EFNB3, heparan sulfate proteoglycans, and the Rac-specific GTPase-activating protein α-chimaerin.22–28 The developing intestine has a similar need for separation between left and right cells and signals, but it seems to accomplish this by a different mechanism—an atypical basement membrane at the midline.

This basement membrane may separate cells that have ingressed from the right and left coelomic epithelia,60 since these cells do not mix at the midline (Fig. 2). It may also prevent mixing of diffusible signals. While the diffusibility of a given signal depends on the tissue context,61 some morphogens (like Nodal) are able to cause effects 200 μm away or more.62 The HH19 DM is only about 150 μm across, not a prohibitive distance for diffusible signals like BMPs, TGFß,40 and CXCL1233 to cross between the left and right DM. Thus, we hypothesized that the DM may need a barrier at the midline to segregate these signals. In support of this, we showed that the midline limits diffusion of dextran from right to left, which suggests that it also blocks the movement of endogenous diffusible signals. Moreover, with a molecular weight of just 3 kDa, dextran’s inability to cross the midline indicates that diffusible proteins of typical weight (such as CXCL12 at 10 kDa and BMP4 at 34 kDa) are also unable to cross. We previously showed evidence that even the 502.78 Da drug AMD3100 was prevented from moving to the other side of the DM.33 While the midline could not be degraded by Ntn4 overexpression (Fig. 5), future studies may reveal tools for the selective destruction of this basement membrane to better understand its function.

This work adds a new facet to our knowledge of basement membrane form and function. Basement membranes play many critical barrier functions in the embryo and adult, usually found as a single layer that underlies polarized epithelial or endothelial cell layers such as those lining the intestines, encircling blood vessels, or enveloping muscle cells, adipocytes, or Schwann cells.46 As such, null mutations in genes encoding basement membrane components often result in embryonic lethality and postnatal pathologies.63–68 However, a role for basement membrane in establishing LR asymmetry has not been described previously.

The basement membrane we describe here is atypical in its double membrane structure, which raises interesting questions about its formation. We showed that the midline is not produced by the mesenchymal cells of the DM (Fig. 4), and that the notochord is not sufficient for its synthesis (Fig. 4). Instead, we consider the endoderm. Upon electroporation of the endoderm with GFP, GFP-positive cells were not detected in the DM later in development but remained restricted to the endoderm (Fig. 4F-H), indicating that the midline does not form from EMT of basement membrane-carrying endodermal cells. We can also eliminate endodermal death as a possible mechanism, because there was no appreciable cell death observed by TUNEL staining at the midline during the stages of interest (data not shown). Nor does the midline form in the same way as other double basement membranes, which arise from the meeting of the basal sides of two tissues.69,70 This is classically illustrated in the kidney glomerulus between epithelial podocytes and endothelial cells70–72 and in the blood-brain barrier between endothelial cells, pericytes, and astrocytes.69,73 Instead in the case of the DM midline barrier, the apical sides of the endodermal cells are facing each other.

We hypothesize that the midline forms when basement membrane is left behind as the endoderm descends ventrally during normal development, as if it were a “scar” of where the endoderm was previously (Fig. 7). During early development the notochord is embedded within the endoderm and a basement membrane covers the two structures. Only later are the two structures separated by a full basement membrane, pointing to a strong connection between the two.74,75 The medial migration of the aortae and coelomic epithelia could provide tissue forces that push the notochord and endoderm apart, leaving behind basement membrane at the DM midline. The observation that the basement membranes of both the endoderm and midline were resistant to disruption by NTN4 could also support the idea that the endoderm is responsible for making both basement membranes (Fig. 7). In support of this, we know that the basement membrane underlying the gut endoderm does not co-migrate with the intestinal epithelial cells as they move from the proliferative intestinal crypts to the tip of the villus over the course of three to six days in the adult; the basement membrane is instead left in place.76 One way to test this hypothesis would be to “trace” endodermal basement membrane by electroporation of a tagged basement membrane component, as has been done recently in C. elegans77–82 and Drosophila.83,84 However, these advances are only just now reaching the mouse system85,86 and have not yet made their way into the chick embryo.

Model of endoderm descending hypothesis for midline formation.

We hypothesize that as the endoderm moves ventrally and the distance between the notochord and endoderm grows, basement membrane from the endoderm may be left behind. This can be compared to a zipper where each side is the basement membrane underlying the endoderm, and when the zipper pull (tip of endoderm) moves downwards, the basement membrane behind it pulls closer together.

The midline barrier is set apart from many other basement membranes by its rapid disappearance. We consider that the loss of the DM midline barrier may be caused in part by “stretching” of the basement membrane as the notochord and endoderm become increasingly separated due to the elongation of the DM (Fig. 3J). Since the DM itself does not contribute basement membrane to the midline (Fig. 4C) and the midline length increases quickly (Fig. 3J), it is plausible that the midline may be pulled until it reaches its ultimate tensile strength (between 0.5-3.8 MPa in other examples of naturally occurring basement membranes),87 and then breaks. This would suggest that midline breakdown is a passive consequence of embryo growth.

However, we also consider that there could be an active breakdown mechanism for the midline. The turnover of stable basement membranes occurs on the scale of weeks,76,88 but the midline barrier appears to degrade over just 12-24 hours. In other contexts, basement membrane destruction can occur over a large area, such as by secretion of matrix metalloproteinases (MMPs) in metastasizing cancer,89,90 or more localized, as exhibited by invadopodia on immune or cancer cells before metastasis.91,92 Such localized basement membrane breakdown is seen in critical developmental processes, including mouth development in deuterostome embryos.93 The oral membrane includes a basement membrane that closes off the digestive system from the outside world. This basement membrane specifically disintegrates to rupture that membrane to open the early mouth cavity.93 In addition, a basement membrane divides the two halves of the embryonic brain, and must be broken down at the site of the corpus callosum for neurons to cross for inter-hemisphere communication in the cerebrum.94,95 Localized basement membrane dissolution is also critical for optic cup fusion, where a contact-dependent dissolution of basement membrane occurs between the two sides of the optic fissure.96,97 Coloboma, a congenital eye defect where some tissue is missing inside the eye resulting in an enlarged, irregular pupil, occurs when optic fusion arrests.98,99 It is worth noting that in all of these examples and the case of the midline, the location of basement membrane breakdown is very specific—neighboring basement membranes appear unaffected (Fig. 3H). Together, these findings support the idea that basement membrane breakdown in the DM may not simply be a passive process but may be required by the embryo for later events in gut development or vascular patterning. Given that a mechanism for basement membrane breakdown is not well characterized in any of the above contexts, there is high potential for future investigation.

The DM midline may have effects on development beyond the intestine. It has long been known that signals produced from axial structures in the embryo are critical for the establishment of LR asymmetry, but it has been difficult for researchers to pin down which “midline structure” or “midline signal” is actually responsible. For example, the heart,100,101 lungs,44 and kidneys102 all rely on “dorsal midline structures” to develop properly. Aberrations from the loss of midline structures like the notochord include “horseshoe kidney,” in which the kidneys stay close to the midline and fuse at their posterior end.103 This is due to the lack of notochordal Shh signaling, but the “midline barrier” that exists downstream from this Shh signaling remains a mystery.104

The DM midline barrier may also play a role in vascular patterning, including aortic fusion. The aorta begins as two parallel tubes with an avascular zone in between (where the midline is),105 but progressively they fuse into one with an anterior-to-posterior wave until the level of the vitelline arteries (Fig. S2A-D).105 Fusion of the aortae coincides with the fragmentation and disappearance of the midline, suggesting that these two processes may be inextricably linked (Fig. 3J, S2). Proper timing of this fusion depends on carefully balanced levels of VEGF,106 SHH,107 and the anterior-to-posterior wave of downregulation of the BMP-inhibitory genes Chordin and Noggin from the notochord.105,108,109 The precise mechanism of dorsal aorta fusion remains unknown, although there is evidence that VEGF signaling pulls VE-cadherin away from its cell-cell junctions.106 This relocalization may be important for the remodeling of the aortic endothelium during fusion and may be linked to the breaking of the DM midline barrier.

Midline structures are critical for laterality to develop correctly. The notochord is certainly involved, particularly as the source of modulators of BMP and Hedgehog signaling. However, this does not seem to be the whole story, given that in each of these contexts the actual “midline barrier” downstream from notochord signals has not been identified. It is possible that the DM midline basement membrane is key here, either for separating left and right signals or perhaps for binding signals67 from the notochord to act as a buffer between each side. The midline barrier may also play a role in the rheology of the DM. Microindentation analyses at HH21 show that the condensed left DM is significantly stiffer than the expanded right DM, and that proper gut tilting is dependent on this difference being tightly regulated.40 It is possible that the midline barrier helps to segregate the stiffness-influencing components of each side (i.e. covalently modified HA on the right,39 N-cadherin on the left36) and provides a “wall” for the right side to push upon in order to swing the gut tube towards the left.

Collectively, we have identified a novel midline barrier in the mesentery that is composed of an atypical double basement membrane that forms a boundary between the left and right sides and limits movement of cells and diffusible signals, independent of Lefty1. The DM midline also presents an opportunity to interrogate the fundamental mechanisms of basement membrane formation and degradation during vertebrate embryonic development, with implications for research on cancer metastasis. We posit that this midline is a distinct strategy for the critical separation of left and right signals and cells, key for establishing and maintaining LR asymmetry for healthy gut development.

Materials and Methods

Chicken embryo development and processing

Fertile chicken eggs were purchased from Westwind Farms (Interlaken, NY http://chickenhawkfood.com). After 36-48 hours of incubation at 37°C, eggs were windowed by removing 8 mL of thin albumen with an 18 ½ gauge needle/10 mL syringe and cutting an oval in the side of the shell, then covering the opening with clear packing tape and returning the egg to the incubator. Once at the desired stage, embryos were isolated in cold 1X PBS and fixed overnight in 2% paraformaldehyde (PFA) at 4°C, followed by PBS washes. Embryos were prepared for cryo-embedding by putting them through graded sucrose solutions ending in 30% sucrose overnight at 4°C. Embryos were cryo-embedded in OCT (VWR 25608-930), sectioned to 15 μm, dried overnight, then stored at −80°C.

Immunofluorescence (IF)

Cryosections were rehydrated in PBS then PBST (0.03% Tween-20), then blocked in 3% heat-inactivated goat serum (HIGS, Gibco 16210072) in PBST for 45 min at room temperature. Primary antibodies were diluted in blocking solution (3% HIGS in PBST) and incubated either for 45 minutes at room temperature or overnight at 4°C. After three PBST washes for five minutes each, secondary antibodies were incubated for 45 minutes at room temperature with 1:2000 dilution of DAPI added. PBST and PBS washes were done before mounting the slides with Prolong Gold anti-fade (Invitrogen P36930). Antigen retrieval pretreatment was necessary for GM130 (BD Biosciences 610822) immunofluorescent staining. Cryosections were rehydrated in water, then microwaved in 1:100 antigen-retrieval solution (Vector Laboratories, H-3300) until nearly boiling. After incubating at 37°C wrapped in aluminum foil for 15 minutes and cooling for 10 minutes, slides were taken through the standard IHC protocol.

RNA in situ hybridization

Section and wholemount RNA in situ hybridization was done using a modified protocol from Moisés Mallo as previously described.110

Dextran injections

Dextran injections into the DM were done using 3000 MW dextran conjugated to fluorescein (ThermoFisher D3306) at a concentration of 10 mg/mL in 1X PBS with Fast Green dye added to better visualize the solution during injection. This mixture was loaded into fine pulled glass capillary needles. A microinjector with a foot pedal was set to 5 psi for 200 ms. Chicken embryos at the desired stage were prepared by removing the vitelline membrane. Injections were done to the right DM only, because embryos lie on their left sides from HH18 onwards and only the right DM is accessible for injection. With the anterior/posterior axis of the embryo perpendicular to the needle and with the needle at a 25° angle, the body wall was gently pulled back so the needle could access the right side of the DM. The needle was gently pressed into the tissue until the embryo moved slightly from the force. Then the foot pedal was pressed once to inject. Embryos were allowed to continue incubating for about two hours, then embryos were collected and fixed in 2% PFA overnight at 4°C. To screen for embryos with quality injections, embryos were cryo-embedded and sectioned. Any embryos with visible damage to the DM in these sections were excluded from further analysis.

CM-DiI and SP-DiO injections

Five μl of a stock solution of CM-DiI or SP-DiO (1μg/μl in EtOH) was diluted into 45 μl of prewarmed 0.3M sucrose in single distilled water maintained at 37°C. Dye solutions were injected into the coelomic epithelium as previously described.44

Electroporation

DM electroporations were performed as described previously.39,111 Endodermal electroporations were performed with a similar method, but with the electroporation mix (plasmid of interest and/or pCAG-GFP, 1X PBS, 1X Fast Green, 1mM MgCl2, and 0.17% carboxymethylcellulose) injected into the empty space beneath the ventral side of the embryo while the negative electrode was held in place there. The positive electrode was placed directly above the negative electrode, centered along the neural tube, before the pulse was applied. pCAGEN-Ntn4 expression plasmid was constructed by cloning the full-length coding sequence out of mouse Ntn4-AP-His plasmid (Addgene 71980) using the primers in Table 1, and cloning into pCAGEN (Addgene 11160) with XhoI (NEB R0146) and NotI (NEB R0189). A second Ntn4 construct was also used, using a mouse Ntn4 (generously provided by Raphael Reuten) and cloned into the pMES vector,112 with similar results.

Notochord transplants

Notochord transplants were performed on HH12-HH15 embryos, using a method adapted from papers describing notochordectomies and notochord transplants.113–119 Note that notochordectomies, while potentially informative, were not done because damage to the endoderm is highly likely in those experiments. Since the endoderm is also potentially implicated in midline formation, an experiment that did not perturb that tissue was preferred. To prepare the donor notochord, the embryo was cut crosswise at the level of the vitelline arteries and close to the end of the tail. Clean 2-4 mm sections of notochord were used for the transplants. A sharp glass needle was used to make an incision in the recipient embryo along the anterior/posterior body axis between the neural tube and somites, deep enough so the ectopic notochord could sit next to the native notochord without puncturing the dorsal aorta. The donor notochord was pressed into place using a pair of dull glass needles or forceps. Embryos continued developing at 37°C until stage HH19.

Scanning electron microscopy (SEM)

Embryos used for Fig. 3C and D were fixed in 4% paraformaldehyde and 2% glutaraldehyde in PBS overnight at 4°C. Samples were then equilibrated in 5% sucrose/PBS for 1 hour at RT, 20% sucrose/PBS for 1 hour at RT, and finally 15% sucrose/7.5% gelatin/PBS at 37 degrees overnight. Embryos were then embedded in plastic molds and frozen in liquid Isopentane in a dry ice-ethanol bath. Cryosections of 10 microns in thickness were collected on poly-L-lysine treated coverslips and incubated twice with fresh drops of PBS for 10 min at 37°C. Coverslips were then washed with 0.1M cacodylate buffer and post fixed with 0.1% osmium tetroxide. Following washes with deuterium-depleted water (DDW), the sections were dehydrated in graded ethanol series, critical point dried (Quorum K850) and sputter coated with 6 nm of chromium (Quorum Q150T). Samples then were viewed on Zeiss Ultra Plus HR Scanning Electron Microscope using the SE2 detector.

Embryos for SEM in Fig. S1 were fixed in 2% glutaraldehyde in 0.50M cacodylate buffer (pH 7.4) at 4°C for two hours, then rinsed three times for 10 minutes each in 0.05M cacodylate buffer. At this point, embryos were cut down to size, using sharp spring scissors to cut the embryo crosswise at the level of the midgut. Then embryos were post-fixed in 1% osmium tetroxide in 0.05M cacodylate buffer at 4°C for one hour, rinsed again in 0.05M cacodylate buffer (3x 10 min), dehydrated in an ethanol series of 25%, 50%, 70%, 95%, and 100% for 10 minutes each, and left in 100% ethanol overnight. The following day embryos were critical point dried in CO2, soaking for 24 hours. Samples were mounted and silver paint was used for conductivity. These embryos were sputter-coated with gold palladium and imaged on a LEO 1550 (Keck SEM).

Imaging, image processing, and quantifications

Brightfield and fluorescent images of tissue sections were taken on a Zeiss Observer Z1 with Apotome, a LSM880 Confocal multiphoton inverted microscope—i880 (Zeiss), or a LSM710 Confocal (Zeiss), or a ScanScope CS2. Stereoscopic images were taken on a SteREO Discovery.V12 (Zeiss). Images were processed using Fiji. Statistical analyses were done using GraphPad Prism.

Quantification of fluorescent intensity of laminin staining was done for five sum intensity projections per biological replicate. For each image, the width of the midline was averaged from three measurements. When no midline is discernible at HH21, the midline width for each image was replaced with the overall average HH20 midline width. Then, five profile plots of gray values (intensity) were obtained from orthogonal lines drawn across the midline. The gray values within the average membrane width centered around the local maximum gray value were averaged to produce a raw mean midline gray value for each image. This process was repeated for neural floor plate basement membrane, and the raw mean midline gray value for each image was normalized to the raw mean floor plate basement membrane value. The means of HH18, HH19, HH20, and HH21 normalized midline gray values were compared via Welch’s unequal variances t-test. Finally, midline length was measured at HH19 as the distance between the tip of the endoderm and the notochord from the hindgut (where the “bird’s nest” of laminin deposition becomes longer than it is wide), to the cranial midgut (where the branches of the dorsal aorta fuse).

GM130 Golgi staining was used to assess cell polarity as described previously.38 Five cell populations were evaluated: cells contacting the midline on the left (“left margin”), cells contacting the midline on the right (“right margin”), right mesenchymal cells not contacting the midline or coelomic epithelium, left coelomic epithelial cells, and rare cells observed between the double membrane of the midline itself (“midline cells”) (Fig. 4D). For each cell, the clockwise angle relative to vertical (0°) of the line drawn between the center of the nucleus and the Golgi apparatus (Fig. 4D inset) was recorded and plotted on an angle histogram in 20° bins with five biological replicates each. The trend of cell polarity in each cell population was assessed using Rayleigh’s test of uniformity using the “circular” R package for circular statistics (https://r-forge.r-project.org/projects/circular/).

eLife Assessment

This study reports the fundamental discovery of a novel structure in the developing gut that acts as a midline barrier between left and right asymmetries. The evidence supporting the dynamics, composition, and function of this novel basement membrane in the chick is in parts solid and in others convincing, but the investigation of its origin and impact on asymmetric organogenesis is not yet conclusive. This careful work is of broad relevance to anyone interested in patterning mechanisms, the importance of the extracellular matrix, and laterality disorders.

Joint Public Review:

When the left-right asymmetry of an animal body is established, a barrier that prevents the mixing of signals or cells across the midline is essential. Such a midline barrier preventing the spreading of asymmetric Nodal signaling during early left-right patterning has been identified. However, midline barriers during later asymmetric organogenesis have remained largely unknown, except in the brain. In this study, the authors discovered an unexpected structure in the midline of the developing midgut in the chick. Using immunofluorescence, they convincingly show the chemical composition of this midline structure as a double basement membrane and its transient existence during the left-right patterning of the dorsal mesentery, which authors showed previously to be essential for forming the gut loop and guiding local vasculogenesis. Labelling experiments suggest a physical barrier function, to cell mixing and signal diffusion in the dorsal mesentery. Cell labelling and graft experiments rule out a cellular composition of the midline from dorsal mesenchyme or endoderm origin and rule out an inducing role by the notochord. Based on laminin expression pattern and Ntn4 resistance, the authors propose a model, whereby the midline basement membrane is progressively deposited by the descending endoderm.

Laterality defects encompass severe malformations of visceral organs, with a heterogenous spectrum that remains poorly understood, by a lack of knowledge of the different players of left-right asymmetry. This fundamental work significantly advances our understanding of left-right asymmetric organogenesis, by identifying an organ-specific and stage-specific midline barrier. The complexities of basement membrane assembly, maintenance, and function are of importance in several other contexts, as for example in the kidney and brain. Thus, this original work is of broad interest.

Overall, reviewers refer to a strong and elegant paper discovering a novel midline structure, combining classic but challenging techniques, to show the dynamics, chemical, and physical properties of the midline. However, reviewers also indicate that further work will be necessary to conclude on the origin and impact of the midline for asymmetric organogenesis. Three issues have been raised to strengthen the claims:

1. The function of the midline as a physical barrier requires clarification. Dextran injection here seems to label cells and not the extracellular space. By counting the proportion of dextran-labeled cells rather than dextran intensity itself, the authors do not measure diffusion per se, but rather cell mixing.

2. The descending endoderm zippering model for the formation of the midline lacks direct evidence. The claim of an endoderm origin is based on laminin expression, but the laminin observed in the midline with an antibody may not necessarily correspond to the same subtype assessed by in situ hybridization. The midline may be Ntn4 resistant until it is injected in the relevant source cells. Alternative origins could be considered, from the bilateral dorsal aortae or the paraxial mesoderm, which would explain the double layer as a meeting point of two lateral tissues.

3. The title implies a role of the midline in left-right asymmetric gut development. However, the importance of the midline is currently inferred from previously published data and stage correlations and will require more direct evidence.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1:

Summary:

Left-right asymmetry in the developing embryo is important for establishing correct lateralisation of the internal organs, including the gut. It has been shown previously that the dorsal mesentery (DM), which supports looping of the endodermal gut tube during development, is asymmetric with sharp delineation of left and right domains prior to gut looping. The authors set out to investigate the nature of the midline barrier that separates the left and right sides of the DM. They identify a transient basement membrane-like structure which is organised into two layers between the notochord and descending endoderm. In the time window when this basement membrane structure exists, there is no diffusion or cell mixing between the left and right sides of the DM, but once this structure starts breaking down, mixing and diffusion occur. This suggests it acts as a barrier, both physical and chemical, between left and right at the onset of gut lateralisation.

Strengths:

The authors identify a new midline structure that likely acts as a barrier to facilitate left and right separation during early organogenesis. This is an interesting addition to the field of laterality, with relevance to laterality-related disorders including heterotaxia, and may represent a gut-specific mechanism for establishing and maintaining early left-right asymmetry. The structure of this midline barrier appears to be an atypical basement membrane, comprising two adjacent basement membranes. The complexities of basement membrane assembly, maintenance, and function are of importance in almost all organismal contexts. Double basement membranes have been previously reported (for example in the kidney glomeruli as the authors note), and increasing evidence suggests that atypical basement membrane organisation or consideration is likely to be more prevalent than previously appreciated. Thus this work is both novel and broadly interesting.

The data presented are well executed, using a variety of well-established methods. The characterisation of the midline barrier at the stages examined is extensive, and the data around the correlation between the presence of the midline barrier and molecular diffusion or cell mixing across the midline are convincing.

Weaknesses:

The study is rather descriptive, and the authors' hypotheses around the origins of the midline barrier are speculative and not experimentally demonstrated. While several potential origins of the midline are excluded raising interesting questions about the timing and cell-type-specific origin of the midline basement membrane, these remain unanswered which limits the scope of the paper.

We extend our appreciation to Reviewer #1 for their thoughtful and comprehensive evaluation of our work, recognizing the considerable time and effort they dedicated to our work. We agree that functional data would significantly strengthen our understanding of the midline barrier and its exact role during LR asymmetric gut development. However, we would like to note that repeated and diligent attempts to perturb this barrier were made using various strategies, such as in vivo laser ablation, diphtheria toxin, molecular disruption (Netrin 4), and enzymatic digestion (MMP2 and MMP9 electroporation) but we observed no significant effect or stable disruption of the midline. We acknowledge and accept this limitation and hope that our discovery will invite future investigations and perturbation of this novel midline structure.

For example, it is unclear whether the two basement membranes originally appear to be part of a single circular/spherical structure (which looks possible from the images) that simply becomes elongated, or whether it is indeed initially two separate basement membranes that extend.

We favor the hypothesis that the elongation of the preexisting small circular structure to an extended double membrane of relatively increased length would be unlikely without continued contribution of new basement membrane components. However, our attempts to label and trace the basement membrane of the endoderm using tagged laminins (LAMB1-GFP, LAMB1-His, and LAMC1-His), and more recently tagged nidogen constructs (NID1-GFP and NID1-mNG) have met with export issues (despite extensive collaboration with experts, Drs. Dave Sherwood and Peter Yurchenco). As such, it remains difficult to differentiate between the two possibilities suggested. We also believe this is an important question and will continue to investigate methods to trace it.

There is a substantial gap between the BMs at earlier stages before the endoderm has descended - is this a lumen, or is it filled with interstitial matrix?

Our preliminary studies indicate that the gap enclosed by the basement membranes in the early midline structure does have extracellular matrix present, such as fibrillin-2 (see Author response image 1). Also, the electron microscopy shown in Fig. 2 C’’ supports that the space between the notochord and endoderm has fibrillar matrix.

The authors show where this basement membrane does not originate from, but only speculate on its origin. Part of this reasoning is due to the lack of Lama1-expressing cells either in the early midline barrier before it extends, or in the DM cells adjacent to it. However, the Laminin observed in the midline could be comprised of a different alpha subtype for example, that wasn't assessed (it has been suggested that the Laminin antibody used in this study is not specific to the alpha-1 subunit, see e.g. Lunde et al, Brain Struct Funct, 2015).

We appreciate this comment and have tried other laminin RNA probes that showed similar lack of midline expression (Lama1, lama3, lama5). Importantly, the laminin alpha 1 subunit is a component of the laminin 111 heterotrimer, which along with laminin 511 is the first laminin to be expressed and assemble in embryonic basement membranes, as reviewed in Yurchenco 2011. Laminin 111 is particularly associated with embryonic development while laminins 511/521 become the most widespread in the adult (reviewed in Aumailley 2013). It is likely that the midline contains laminin 111 based on our antibody staining and the accepted importance and prevalence of laminin 111 in embryonic development. However, it is indeed worth noting that most laminin heterotrimers contain beta 1, gamma 1, or both subunits, and due to this immunological relation laminin antibody cross reactivity is certainly known (Aumailley 2013). As such, while laminin 511 remains a possibility as a component of the midline BM, our lama5 in situs have shown no differential expression at the midline of the dorsal mesentery (see Author response image 2), and as such we are confident that our finding of no local laminin transcription is accurate. Additionally, we will note that the study referenced by the Reviewer observed cross reactivity between the alpha 1 and alpha 2 subunits. Laminin 211/221 is an unlikely candidate based on the embryonic context, and because they are primarily associated with muscle basement membranes (Aumailley 2013). In further support, we recently conducted a preliminary transcriptional profile analysis of midline cells isolated through laser capture microdissection (LCM), which revealed no differential expression of any laminin subunit at the midline. Please note that these data will be included as part of a follow-up story and falls beyond the scope of our initial characterization.

Similarly, the authors show that the midline barrier breaks down, and speculate that this is due to the activity of e.g. matrix metalloproteinases, but don't assess MMP expression in that region.

This is an important point, as the breakdown of the midline is unusually rapid. Our MMP2 RNA in situ hybridization at HH21, and ADAMTS1 (and TS9) at HH19-21 indicates no differential activity at the midline (see Author response images 3 and 4). Our future focus will be on identifying a potential protease that exhibits differential activity at the midline of the DM.

The authors suggest the (plausible) hypothesis that the descent of the endoderm pulls or stretches the midline barrier out from its position adjacent to the notochord. This is an interesting possibility, but there is no experimental evidence to directly support this. Similarly, while the data supporting the barrier function of this midline is good, there is no analysis of the impact of midline/basement membrane disruption demonstrating that it is required for asymmetric gut morphogenesis. A more functional approach to investigating the origins and role of this novel midline barrier would strengthen the study.

Yes, we fully agree that incorporating functional data would immensely advance our understanding of the midline barrier and its crucial role in left-right gut asymmetry. However, our numerous efforts to perturb this barrier have encountered technical obstacles. For instance, while perturbing the left and right compartments of the DM is a routine and well-established procedure in our laboratory, accessing the midline directly through similar approaches has been far more challenging. We have made several attempts to address this hurdle using various strategies, such as in vivo laser ablation, diphtheria toxin, molecular disruption (Netrin 4), and enzymatic digestion (MMP2 and MMP9 electroporation). Despite employing diverse approaches, we have yet to achieve effective and interpretable perturbation of this resilient structure. We acknowledge this limitation and remain committed to developing methods to disrupt the midline in our current investigations. We again thank Reviewer #1 for the detailed feedback on our manuscript, guidance, and the time taken to provide these comments.

Recommendations For The Authors:

Using Laminin subunit-specific antibodies, or exploring the mRNA expression of more laminin subunits may support the argument that the midline does not derive from the notochord, endoderm, or DM.

As mentioned above, RNA in situ hybridization for candidate genes and a preliminary RNA-seq analysis of cells isolated from the dorsal mesentery midline revealed no differential expression of any laminin subunits.

Similarly, expression analysis of Laminin-degrading MMPs, and/or application of an MMP inhibitor and assessment of midline integrity could strengthen the authors' hypothesis that the BM is actively and specifically broken down.

Our MMP2 RNA in situ hybridization at HH21, and ADAMTS1 at HH19-21shows no differential expression pattern at the midline of the DM (see Author response image 3). We have not included these data in the revision, but future work on this topic will aim at identifying a protease that is differentially active at the midline of the DM.

Functionally testing the role of barrier formation in regulating left-right asymmetry or the role of endoderm descent in elongating the midline barrier would be beneficial. Regarding the former, the authors show that Netrin4 overexpression is insufficient to disrupt the midline, but perhaps overexpression of e.g. MMP9 prior to descent of the endoderm would facilitate early degradation of the midline, and the impact of this on gut rotation could be assessed.

Unfortunately, MMP9 electroporation has produced little appreciable effect. We acknowledge that the lack of direct evidence for the midline’s role in regulating left-right asymmetry is a shortcoming, but current work on this subject aims to define the midline’s function to LR asymmetric morphogenesis.

Reviewer #2:

When the left-right asymmetry of an animal body is established, the barrier that prevents the mixing of signals or cells across the midline is essential. The midline barrier that prevents the mixing of asymmetric signals during the patterning step has been identified. However, a midline barrier that separates both sides during asymmetric organogenesis is unknown. In this study, the authors discovered the cellular structure that seems to correspond to the midline in the developing midgut. This midline structure is transient, present at the stage when the barrier would be required, and composed of Laminin-positive membrane. Stage-dependent diffusion of dextran across the midline (Figure 6) coincides with the presence or absence of the structure (Figures 2, 3). These lines of indirect evidence suggest that this structure most likely functions as the midline barrier in the developing gut.

We extend our gratitude to Reviewer #2 for their thoughtful assessment of our research and for taking the time to provide these constructive comments. We are excited to report that we have now included additional new data on midline diffusion using BODIPY and quantification method to further support our findings on the midline's barrier function. While our data on dextran and now BODIPY both indirectly suggests barrier function, we aspire to perturb the midline directly to assess its role in the dorsal mesentery more conclusively. However, our numerous efforts to perturb this barrier have encountered technical obstacles. For instance, while perturbing the left and right compartments of the DM is a routine and well-established procedure in our laboratory, accessing the midline directly through similar approaches has been far more challenging. We have made several attempts to address this hurdle using various strategies, such as in vivo laser ablation, diphtheria toxin, molecular disruption (Netrin 4), and enzymatic digestion (MMP2 and MMP9 electroporation). Despite employing diverse approaches, we have yet to achieve effective and interpretable perturbation of this resilient structure. Moving forward, our focus is on identifying an effective means of perturbation that can offer direct evidence of barrier function.

Recommendations For The Authors:

(1) It would be much nicer if the requirement of this structure for asymmetric morphogenesis was directly tested. However, experimental manipulations such as ectopic expression of Netrin4 or transplantation of the notochord were not able to influence the formation of this structure (these results, however, suggested the mechanism of the midline formation in the gut dorsal mesentery). Therefore, it seems not feasible to directly test the function of the structure, and this should be the next issue.

We fully agree that the midline will need to be perturbed to fully elucidate its role in asymmetric gut morphogenesis. As noted, multiple attempts were ineffective at perturbing this structure. Extensive current work on this topic is dedicated to finding an effective perturbation method.

(2) Whereas Laminin protein was present in the double basement membrane at the midline, Laminin mRNA was not expressed in the corresponding region (Fig. 4A-C). It is necessary to discuss (with experimental evidence if available) the origin of Laminin protein.

As we have noted, the source of laminin and basement membrane components for the midline remains unclear - no local transcription and the lack of sufficiency of the notochord to produce a midline indicates that the endoderm to be a likely source of laminin, as we have proposed in our zippering endoderm model. We will note that Fig. 4A-C indicate that laminin is in fact actively transcribed in the endoderm. Currently, attempts to trace the endodermal basement membrane using tagged laminins (LAMB1-GFP, LAMB1-His, and LAMC1-His), and more recently tagged nidogen constructs (NID1-GFP and NID1-mNG) have met with export issues (despite extensive collaboration with experts, Drs. Dave Sherwood and Peter Yurchenco). Confirmation of our proposed endodermal origin model is a goal of our ongoing work.

(3) Figure 4 (cell polarity from GM130 staining): addition of representative GM130 staining images for each Rose graph (Figure 4E) would help. They can be shown in Supplementary Figures. Also, a graph for the right coelomic epithelium in Fig. 4E would be informative.

We have added the requested GM130 images in our Supplemental Figures (please refer to Fig. S4ABB’) and modified the main Fig. 4E to include a rose graph for the polarity of the right coelomic epithelium.

(4) Histological image of HH19 DM shown in Fig. 2J looks somehow different from that shown in Fig. 3F. Does Fig. 2J represent a slightly earlier stage than Fig. 3F?

Figure 2J and Figure 3F depict a similar stage, although the slight variation in the length of the dorsal mesentery is attributed to the pseudo time phenomenon illustrated in Figure 3J-J’’’. This implies that the sections in Figure 2J and Figure 3F might originate from slightly different positions along the anteroposterior axis. Nonetheless, these distinctions are minimal, and based on the dorsal mesentery's length in Figure 2J, the midline is likely extremely robust regardless of this minor pseudo time difference.

Reviewer #3:

Summary:

The authors report the presence of a previously unidentified atypical double basement membrane (BM) at the midline of the dorsal mesentery (DM) during the establishment of left-right (LR) asymmetry. The authors suggest that this BM functions as a physical barrier between the left and the right sides of the DM preventing cell mixing and ligand diffusion, thereby establishing LR asymmetry.

Strengths:

The observation of the various components in the BM at the DM midline is clear and convincing. The pieces of evidence ruling out the roles of DM and the notochord in the origin of this BM are also convincing. The representation of the figures and the writing is clear.

Weaknesses:

The paper's main and most important weakness is that it lacks direct evidence for the midline BM's barrier and DM LR asymmetry functions.

We thank Reviewer #3 for their thoughtful and comprehensive evaluation of our work, recognizing the considerable time and effort they dedicated to assessing our study. We fully agree that incorporating functional data would immensely advance our understanding of the midline barrier and its crucial role in left-right gut asymmetry. However, several distinct attempts at perturbing this barrier have encountered technical obstacles. While our laboratory routinely perturbs the left and right compartments of the DM via DNA electroporation and other techniques, directly perturbing the midline using these methods is far more challenging. We have made diligent attempts to address this using various strategies, such as in vivo laser ablation, diphtheria toxin, molecular disruption (Netrin 4), and enzymatic digestion (MMP2 and MMP9 electroporation). However, we have not yet been able to identify a means of producing consistent and interpretable perturbation of the midline. We acknowledge this limitation and remain committed to developing methods to disrupt the midline in our current investigations.

Recommendations For The Authors:

Major:

(1) We suggest the authors test their hypotheses i.e., physical barrier and proper LR asymmetry establishment by the midline BM, by disrupting it using techniques such as physical ablation, over-expression of MMPs, or treatment with commercially available enzymes that digest the BM.

As above, efforts involving physical ablation and MMP overexpression have not yielded significant effects on the midline thus far. Moving forward, investigating the midline's role in asymmetric morphogenesis will necessitate finding a method to perturb it effectively. In pursuit of progress on this critical question, we recently conducted laser capture microdissection (LCM) and RNA-sequencing of the midline to unravel the mechanisms underlying its formation and potential disruption. This work shows promise but it is still in its early stages; validating it will require significant time and effort, and it falls outside the scope of the current manuscript.

(2) Lefty1's role in the midline BM was ruled out by correlating lack of expression of the gene at the midline during HH19 when BM proteins expression was observed. Lefty1 may still indirectly or directly trigger the expression of these BM proteins at earlier stages. The only way to test this is by inhibiting lefty1 expression and examining the effect on BM protein localization.

We have added a section to discuss the potential of Lefty1 inhibition as a future direction. However, similar to perturbing global Nodal expression, interpreting the results of Lefty1 inhibition could be challenging. This is because it may not specifically target the midline but could affect vertebrate laterality as a whole. Despite this complexity, we acknowledge the value of such an experiment and consider it worth pursuing in the future.

(3) Using a small dextran-based assay, the authors conclude that diffusible ligands such as cxcl2 and bmp4 do not diffuse across the midline (Figure 6). However, dextran injection in this system seems to label the cells, not the extracellular space. The authors measure diffusion, or the lack thereof, by counting the proportion of dextran-labeled cells rather than dextran intensity itself. Therefore, This result shows a lack of cell mixing across the midline (already shown in Figure 2 ) rather than a lack of diffusion.

We should emphasize that the dextran-injected embryos shown in Fig. 6 D-F were isolated two hours post-injection, a timeframe insufficient for cell migration to occur across the DM (Mahadevan et al., 2014). We also collected additional post-midline stage embryos ten minutes after dextran injections - too short a timeframe for significant cellular migration (Mahadevan et al., 2014). Importantly, the fluorescent signal in those embryos was comparable to that observed in the embryos in Fig. 6. Thus, we believe the movement of fluorescent signal across the DM when the barrier starts to fragment (HH20-HH23) is unlikely to represent cell migration. More than a decade of DNA electroporation experiments of the left vs. right DM by our laboratory and others have never indicated substantial cell migration across the midline (Davis et al., 2008; Kurpios et al., 2008; Welsh et al., 2013; Mahadevan et al., 2014; Arraf et al. 2016; Sivakumar et al., 2018; Arraf et al. 2020; and Sanketi et al., 2022). This is also shown in our current GFP/RFP double electroporation data in Fig. 2 G-H, and DiI/DiO labeling data in Fig. 2 E-G. Collectively, our experiments suggest that the dextran signal we observed at HH20 and HH23 is likely not driven by cell mixing.

To further strengthen this argument, we now have additional new data on midline diffusion using BODIPY diffusion and quantification method to support our findings on the midline's function against diffusion (please refer to New Fig. 6H-M). Briefly, we utilized a BODIPY-tagged version of AMD3100 (Poty et al., 2015) delivered via soaked resin beads surgically inserted into the left coelomic cavity (precursor to the DM). The ratio of average AMD3100-BODIPY intensity in the right DM versus the left DM was below 0.5 when the midline is intact (HH19), indicating little diffusion across the DM (Fig. 6J). At HH21 when no midline remains, this ratio significantly rises to near one, indicating diffusion of the drug is not impeded when the midline basement membrane structure is absent. Collectively, these data suggest that the basement membrane structure at the midline forms a transient functional barrier against diffusion.

(4) Moreover, in a previous study (Mahadevan et al., Dev Cell., 2014), cxcl2 and bmp4 expression was observed on both the left and right side before gut closure (HH17, when midline BM is observed). Then their expression patterns were restricted on the left or right side of DM at around HH19-20 (when midline BM is dissociated). The authors must explain how the midline BM can act as a barrier against diffusible signals at HH-17 to 19, where diffusible signals (cxcl12 and bmp4) were localized on both sides.

We appreciate the Reviewer's invitation to clarify this crucial point. Early in dorsal mesentery (DM) formation, genes like Cxcl12 (Mahadevan et al., Dev Cell 2014) and Bmp4 (Sanketi et al., Science 2021) exhibit symmetry before Pitx2 expression initiates on the left (around ~HH18, Sanketi et al., 2021). Pitx2 then inhibits BMP4 (transcription) and maintains Cxcl12 (mRNA) expression on the left side. The loss of Cxcl12 mRNA on the right is due to the extracellular matrix (ECM), particularly hyaluronan (Sivakumar et al., Dev Cell 2018). Our hypothesis is that during these critical stages of initial DM asymmetry establishment, the midline serves as a physical barrier against protein diffusion to protect this asymmetry during a critical period of symmetry breaking. Although some genes, such as Pitx2 and Cxcl12 continue to display asymmetric transcription after midline dissolution (Cxcl12 becomes very dynamic later on – see Mahadevan), it's crucial to note that the midline's primary role is preventing protein diffusion across it, akin to an insurance policy. Thus, the absence of the midline barrier at HH21 does not result in the loss of asymmetric mRNA expression. We think its primary function is to block diffusible factors from crossing the midline at a critical period of symmetry breaking. We acknowledge that confirming this hypothesis will necessitate experimental disruption of the midline and observing the consequent effects on asymmetry in the DM. This remains central to our ongoing research on this subject.

(5) On page 11, lines 15-17, the authors mention that "We know that experimentally mixing left and right signals is detrimental to gut tilting and vascular patterning-for example, ectopic expression of pro-angiogenic Cxcl12 on the right-side results in an aberrant vessel forming on the right (Mahadevan et al., Dev Cell., 2014)". In this previous report from the author's laboratory, the authors suggested that ectopic expression of cxcl12 on the right side induced aberrant formation of the vessel on the right side, which was formed from stage HH17, and the authors also suggested that the vessel originated from left-sided endothelial cells. If the midline BM acts as a barrier against the diffusible signal, how the left-sided endothelial cells can contribute to vessel formation at HH17 (before midline BM dissociation)?

To address this point, we suggest directing the Reviewer to previously published supplemental movies of time-lapse imaging, which clearly illustrate the migration path of endothelial cells from left to right DM (Mahadevan et al., Dev Cell 2014). While the Reviewer correctly notes that ectopic induction of Cxcl12 on the right induces left-to-right migration, it's crucial to highlight that these cells never cross the midline. Instead, they migrate immediately adjacent to the tip of the endoderm (please also refer to published Movies S2 and S3). We observe this migration pattern even in wild-type scenarios during the loss of the endogenous right-sided endothelial cords, where some endothelial cells from the right begin slipping over to the left around HH19-20 (over the endoderm), as the midline is beginning to fragment, but never traverse the midline. We attribute this migration pattern to a dorsal-to-ventral gradient of left-sided Cxcl12 expression, as disrupting this pattern perturbs the migration trajectory (Mahadevan).

1. It is unclear how continuous is the midline BM across the anterior-posterior axis across the relevant stages. Relatedly, it is unclear how LR segregated the cells are, across the anterior-posterior axis across the relevant stages.

We refer the reviewer to Fig. 3J-K, in which the linear elongation of the midline basement membrane structure is shown and measured at HH19 in three embryos from the posterior of the embryo to the anterior point at which the midline is fragmented and ceases to be continuous. Similarly, Fig. S2 shoes the same phenomenon in serial sections along the length of the anterior-posterior (AP) axis at HH17, also showing the continuity of the midline. All our past work at all observed sections of the AP axis has shown that cells do not move across the midline as indicated by electroporation of DNA encoding fluorescent reporters (Davis et al. 2008, Kurpios et al. 2008, Welsh et al. 2013, Mahadevan et al. 2014, Sivakumar et al. 2018, Sanketi et al. 2022), and is shown again in Fig. 2 E-H. As noted previously, very few endothelial cells cross the midline at a point just above the endoderm (image above) when the right endothelial cord remodels (Mahadevan et al. 2014), but this is a limited phenomenon to endothelial cells and cells of the left and right DM are fully segregated as previously established.

Minor comments:

(1) The authors found that left and right-side cells were not mixed with each other even after the dissociation of the DM midline at HH21 (Fig2 H). And the authors also previously mentioned that N-cadherin contributes to cell sorting for left-right DM segregation (Kurpios et al., Proc Natl Acad Sci USA., 2008). It could be a part of the discussion about the difference in tissue segregation systems before or after the dissociation of DM midline.

We appreciate this thoughtful suggestion. N-cadherin mediated cell sorting is key to the LR asymmetry of the DM and gut tilting, and we believe it underlies the observed lack of cell mixing from left and right DM compartments after the midline fragments. We have added a brief section to the discussion concerning the asymmetries in N-cadherin expression that develop after the midline fragments.

(2) Please add the time point on the images (Fig3 C, D, Fig 6A and B)

We have updated these figures to provide the requested stage information.

(3) The authors suggested that the endoderm might be responsible for making the DM BM midline because the endoderm links to DM midlines and have the same resistance to NTN4. The authors mentioned that the midline and endoderm might have basement membranes of the same "flavor." However, perlecan expression was strongly expressed in the midline BM compared with the endodermal BM. It could be a part of the discussion about the difference in the properties of the BM between the endoderm and DM midline.

Perlecan does indeed localize strongly to the endoderm as well as the midline. The HH18 image included in prior Fig. S3 B’, B’’ appears to show atypically low antibody staining in the endoderm for all membrane components. Perlecan is an important component for general basement membrane assembly, and the bulk of our HH18 and HH19 images indicate strong staining for perlecan in both midline and endoderm. Perlecan staining at the very earliest stages of midline formation also indicate perlecan in the endoderm as well, supporting the endoderm as a potential source for the midline basement membrane. We have updated Fig. S3 to include these images in our revision.

(4) The authors investigated whether the midline BM originates from the notochord or endoderm, but did not examine a role for endothelial cells and pericytes surrounding the dorsal aorta (DA). In Fig S1, Fig S2, and FigS3, the authors showed that DA is very close to the DM midline basement membrane, so it is worth checking their roles.

We fully agree that the dorsal aorta and the endothelial cords that originate from the dorsal aorta may interact with the midline in important ways. However, accessing the dorsal aorta for electroporation or other perturbation is extremely difficult. Additionally, the basement membrane of vascular endothelial cells has a distinct composition from a non-vascular basement membrane. Vascular endothelial cells produce only alpha 4 and alpha 5 laminin subunits but contain no alpha 1 subunit in any known species (reviewed in DiRusso et al., 2017). Thus, endothelial cell-derived basement membranes would not contain the alpha 1 laminin subunit that we used in our studies as a robust marker of the midline basement membrane. Additionally, no fibronectin is found in the midline basement membrane, while it is enriched in the dorsal aorta (see Supplemental Figure 3CC’C’’). We will briefly note that our preliminary data in quail tissue indicates that QH1+ cord cells (i.e. endothelial cells) sometimes exhibit striking contact with the midline along the dorso-ventral length of the DM, suggesting not an origin but an important interaction.

Reviewer #4 (Recommendations For The Authors):

Major comments:

(1) The descending endoderm zippering model for the formation of the midline lacks evidence.

We have attempted to address this issue by introducing several tagged laminin constructs (LAMB1-GFP, LAMB1-His, LAMC1-His), and more recently tagged nidogen plasmids (NID1-GFP and NID1-mNG) to the endoderm via DNA electroporation to try to label the source of the basement membrane. Production of the tagged components occurred but no export was observed in any case (despite extensive collaboration with experts in this area, Drs. Dave Sherwood and Peter Yurchenco). This experiment was further complicated by the necessary large size of these constructs at 10-11kb due to the size of laminin subunit genes, resulting in low electroporation efficiency. We also believe this is an important question and are continuing to investigate methods to trace it.

The midline may be Ntn4 resistant until it is injected in the source cells.

Ntn4 has been shown to disrupt both assembling and existing basement membranes (Reuten et al. 2016). Thus, we feel that the midline and endodermal basement membranes’ resistance to degradation is not determined by stage of assembly or location of secretion.

Have you considered an alternative origin from the bilateral dorsal aorta or the paraxial mesoderm, which would explain the double layer as a meeting of two lateral tissues? The left and right paraxial mesoderm seem to abut in Fig. S1B-C and S2E, and is laminin-positive in Fig 4A'. What are the cells present at the midline (Fig.4D-E)? Are they negative for the coelomic tracing, paraxial or aortic markers?

We fully agree that alternate origins of the midline basement membrane cannot be ruled out from our existing data. We agree and have considered the dorsal aorta and even the endothelial cords that originate from the dorsal aorta. However, accessing the dorsal aorta for electroporation or other perturbation is extremely difficult. Importantly, the basement membrane of vascular endothelial cells has a distinct composition from a non-vascular basement membrane. Vascular endothelial cells produce only alpha 4 and alpha 5 laminin subunits but contain no alpha 1 subunit in any known species (reviewed in Hallmann et al. 2005). Thus, endothelial cell-derived basement membranes would not contain the alpha 1 laminin subunit that we used in our studies as a robust marker of the midline basement membrane. Note in Fig. 3 E-H that our laminin alpha 1 antibody staining does not label the aortae. Additionally, no fibronectin is found in the midline basement membrane, while it is enriched in the dorsal aorta (see Supplemental Figure 3CC’C’’). We will briefly note that our preliminary data in quail tissue indicates that QH1+ cord cells (i.e. endothelial cells) sometimes exhibit striking contact with the midline along the dorso-ventral length of the DM, suggesting not an origin but an important interaction. Moreover, at the earliest stages of midline basement membrane emergence, the dorsal aortae are distant from the nascent basement membrane, as are the somites, which have not yet undergone any epithelial to mesenchymal transition. Fig. S2G provides an example of an extremely early midline basement membrane without dorsal aorta or somite contact. S2G is from a section of the embryo that is fairly posterior in the embryo, it is thus less developed in pseudo-time and gives a window on midline formation in very early embryos.

(2) The importance of the midline is inferred from previously published data and stage correlations but will require more direct evidence. Can the midline be manipulated with Hh signaling or MMPs?

We agree that direct evidence in the form of midline perturbation will be critically required. As previously noted, our numerous efforts to perturb this barrier have encountered technical obstacles. For instance, while perturbing the left and right compartments of the DM is a routine and well-established procedure in our laboratory, accessing the midline directly through similar approaches has been far more challenging. We have made several attempts to address this hurdle using various strategies, such as in vivo laser ablation, diphtheria toxin, molecular disruption (Netrin 4), and enzymatic digestion (MMP2 and MMP9 electroporation). Despite employing diverse approaches, we have yet to achieve effective and interpretable perturbation of this resilient structure. Targeting Hh signaling between the endoderm and notochord is a good idea and we will continue these efforts. Thanks very much.

Minor comments:

- Please add the species in the title.

We have altered the title as follows: “An atypical basement membrane forms a midline barrier during left-right asymmetric gut development in the chicken embryo.”

- The number of observations in Fig2, Fig3A-B, 4A-C, G-H, S1, S3 is lacking.

We have added the requested n numbers of biological replicates to the legends of the specified figures.

- Please annotate Fig 3J to show what is measured in K.

We have modified Fig. 3J to include a dashed bar indicating the length measurements in Fig. 3K.

- Please provide illustrations of Fig 4E.

We have added a representative image of GM130 staining to the supplement.

- If laminin gamma is the target of Ntn4, its staining would help interpret the results of Ntn4 manipulation. Is laminin gamma present in different proportions in the different types of basement membranes, underlying variations in sensitivity?

Laminin is exported as a heterotrimer consisting of an alpha, beta, and gamma subunit. Laminin gamma is therefore present in equal proportions to other laminins in all basement membranes with a laminin network. Several gamma isoforms do exist, but only laminin gamma 1 will bind to laminin alpha 1, which we use throughout this paper to mark the midline as well as nearby basement membranes that are sensitive to Ntn4 disruption. Thus, gamma laminin proportions or isoforms are unlikely to underlie the resistance of the midline and endodermal basement membranes to Ntn4 (reviewed in Yurchenco 2011).

- Please comment: what is the red outline abutting the electroporated DM on the left of Fig5B?

The noted structure is the basement membrane of the nephric duct – we added this information to Fig. 5B image and legend.

- The stage in Fig 6A-B is lacking.

We have added the requested stage information to Fig. 6.

- Please comment on whether there is or is not some cell mixing Fig 2H, at HH21 after the midline disappearance. Is it consistent with Fig. 6E-F which labels cells?

More than a decade of DNA electroporation experiments of the left vs. right DM by our laboratory and others have never indicated dorsal mesentery cell migration across the midline (Davis et al., 2008; Kurpios et al., 2008; Welsh et al., 2013; Mahadevan et al., 2014; Arraf et al. 2016; Sivakumar et al., 2018; Arraf et al. 2020; and Sanketi et al., 2022). This is also shown in our current GFP/RFP double electroporation data in Fig. 2 G-H, and DiI/DiO labeling data in Fig. 2 E-G. Cell mixing does not occur even after midline disappearance, most likely due to asymmetric N-cadherin expression on the left side of the DM (Kurpios et al., 2008). The sparse, green-labeled cells observed on the right side in Fig. 2H are likely a result of DNA electroporation - the accuracy of this process relies on the precise injection of the left (or right) coelomic cavity (precursor to the gut mesenchyme including the DM) and subsequent correct placement of the platinum electrodes.

Based on these data, we strongly feel that cellular migration is not responsible for the pattern of dextran observed in Fig. 6E-F, especially in light of the N-cadherin mediated segregation of left and right. We will also note that there is no significant difference between dextran diffusion at HH19 and HH20, only a trend towards significance. Additionally, we would like to note that the dextran-injected embryos were isolated two hours post-injection, which we do not believe is sufficient time for any cell migration to occur across the DM. We also collected additional post-midline stage embryos ten minutes after dextran injections (data not shown), too short a timeframe for significant cellular migration, and the fluorescent signal in those embryos was comparable to that represented in the embryos in Fig. 6. Thus, we believe the movement of fluorescent signal across the DM observed when the barrier starts to fragment at HH20 and HH23 is unlikely to represent movement of cells.

To further strengthen this argument, we now have additional new data on midline diffusion using BODIPY and quantification method to support our findings on the midline's function against diffusion (please refer to New Fig. 6H-M). Briefly, we utilized a BODIPY-tagged version of AMD3100 (Poty et al., 2015) delivered via soaked resin beads surgically inserted into the left coelomic cavity (precursor to the DM). The ratio of average AMD3100-BODIPY intensity in the right DM versus the left DM was below 0.5 when the midline is intact (HH19), indicating little diffusion across the DM (Fig. 6J). At HH21 when no midline remains, this ratio significantly rises to near one, indicating diffusion of the drug is not impeded when the midline basement membrane structure is absent. Collectively, these data suggest that the basement membrane structure at the midline forms a transient functional barrier against diffusion.

- 'independent of Lefty1': rephrase or show the midline phenotype after lefty1 inactivation.

We agree with this comment and have rephrased this section to indicate the midline is present “at a stage when Lefty1 is no longer expressed at the midline.”

We again would like to extend our sincere gratitude to our reviewers and the editors at eLife for their dedicated time and thorough evaluation of our paper. Their meticulous attention to detail and valuable insights have strengthened our data and provided further support for our findings.

Version 2

Abstract

Abstract

Correct intestinal morphogenesis depends on the early embryonic process of gut rotation, an evolutionarily conserved program in which a straight gut tube elongates and forms into its first loops. However, the gut tube requires guidance to loop in a reproducible manner. The dorsal mesentery (DM) connects the gut tube to the body and directs the lengthening gut into stereotypical loops via left-right (LR) asymmetric cellular and extracellular behavior. The LR asymmetry of the DM also governs blood and lymphatic vessel formation for the digestive tract, which is essential for prenatal organ development and postnatal vital functions including nutrient absorption. Although the genetic LR asymmetry of the DM has been extensively studied, a divider between the left and right DM has yet to be identified. Setting up LR asymmetry for the entire body requires a Lefty1+ midline barrier to separate the two sides of the embryo, without it, embryos have lethal or congenital LR patterning defects. Individual organs including the brain, heart, and gut also have LR asymmetry, and while the consequences of left and right signals mixing are severe or even lethal, organ-specific mechanisms for separating these signals are poorly understood. Here, we uncover a midline structure composed of a transient double basement membrane, which separates the left and right halves of the embryonic chick DM during the establishment of intestinal and vascular asymmetries. Unlike other basement membranes of the DM, the midline is resistant to disruption by intercalation of Netrin4 (Ntn4). We propose that this atypical midline forms the boundary between left and right sides and functions as a barrier necessary to establish and protect organ asymmetry.

Introduction

Humans and most other animals are bilaterians—animals whose left and right external features can be mirrored—but often the internal organs have striking LR asymmetries. For example, in humans the heart resides on the left side of the thoracic cavity, the liver is predominantly on the right, and the spleen is on the left. Even paired organs like the lungs can show LR asymmetries— the left human lung has two lobes while the right lung has three. The left and right sides of the body are specified early in development, after the anterior/posterior and dorsal/ventral axes have been established.1,2 This patterning relies heavily on the expression of Sonic hedgehog (Shh) upstream of Nodal on the left side, and the repression of these genes by Activin on the right side.3 The localization of SHH-producing cells to the left is accomplished by nodal flow (mouse,4,5 zebrafish6 and Xenopus7 embryos) or rotational cell movements around the node (chicken and pig embryos),8 as well as cell death at the embryonic midline that may be a consequence of its abundant extracellular matrix (ECM).9

Establishing the vertebrate LR body axis depends on a midline barrier to separate side-specific diffusible signals.10–15 This is achieved with a highly specific expression pattern of an inhibitor, Lefty1, at the center of the embryo which prevents the diffusible, left-sided signal NODAL from crossing to the embryo’s right side.11–15 Sixty percent of Lefty1-knockout mouse embryos die in utero and an additional 20% die before weaning, suffering from left isomerism of the lungs (in other words, both lungs have left lung lobation) and positional defects of the heart and the major vessels leading into/out of it.13 Other important examples of laterality defects are seen in conjoined twins, in which an embryo divides partially at the primitive streak stage (Day 13-14 of gestation for humans).16 In laterally conjoined (dicephalus) twins which form when the two primitive streaks are parallel, the left side of one embryo and the right side of the other are connected without the LEFTY1 barrier in between. Consequently, these conjoined twins often exhibit LR defects.17,18

This underscores that separation of left and right signals is fundamental in early development for setting up correct placement and LR asymmetric patterning of individual organs including the heart and gut.6,10,19–21 Of all the organs with LR asymmetry, only the brain is known to harbor an organ-specific midline barrier.22–28 Thus, midline barriers may have broad developmental significance for the embryo and its organs, yet very few such structures have been characterized. Even organs that do not have obvious LR differences in the adult develop as a result of conserved LR asymmetric morphogenesis. The intestine is an excellent model for this, especially given the relatively simple tubular structure of the organ itself. During development, the intestine grows to great lengths (about eight meters in adult humans),29 and this long tube must be looped to fit into the body cavity in a stereotypical, species-specific way.30,31 When the developmental program directing gut looping is perturbed, as is the case for one in 500 infants who have congenital malrotation of the gut, there is an increased risk of volvulus, a lethal self-strangulation of the gut that requires immediate pediatric surgical intervention.32

The gut tube does not loop autonomously. Rather, gut looping is directed by the neighboring dorsal mesentery (DM) (Fig. 1A), a thin mesodermal organ that connects the gut to the rest of the body and through which intestinal blood and lymphatic vessels traverse.31,33–35 The left and right sides of the DM take on different properties at the molecular, cellular, and extracellular levels which are critical to initiate asymmetric gut looping and vascular morphogenesis.33,36–40 Gut tilting is the symmetry-breaking event that initiates asymmetric gut looping, which occurs at embryonic day 10.5 (E10.5) in the mouse and Hamburger-Hamilton stage 19-21 (HH19-21)41 in the chicken embryo (Fig. 1B).37 Gut tilting is driven by the condensation of the ECM in the left DM, and an expansion of the ECM on the right (Fig. 1B).36 Blood vasculature also develops asymmetrically. Although endothelial precursors exist on both the left and right DM prior to tilting, as the asymmetries are established the right-sided endothelial cells emigrate rather than forming vessels.33 Only a left-sided gut artery develops further (Fig. 1B), which goes on to supply blood to a significant portion of the adult intestine (Fig. 1A).33 Interestingly, these right-sided endothelial cells emigrate dorsally and ventrally, but they cannot cross directly over to the left side.33 This might indicate the presence of a barrier against cell migration at the midline.

LR asymmetry in the DM is critical for proper gut looping and vascular patterning.

(A) Asymmetries in the DM drive the formation of the first and subsequent gut loops. Concurrently, the vasculature is being patterned in the DM. The 1° longitudinal artery (*) gives rise to the ileocolic artery, which provides a significant portion of the adult intestine with critical blood flow. (B) HH17: The DM has cellular symmetry. HH18 and 19: The right mesenchyme begins expanding and the right epithelial cells elongate. The right-sided endothelial cells (vascular precursors) begin to disperse and leave the compartment, while the left-sided endothelium is maintained to become the future gut arteries. HH22: The asymmetric forces have pushed the gut tube to the left. A left-sided blood vessel, the 1° longitudinal artery (*) has formed. GT = gut tube, DM = dorsal mesentery, DA = dorsal aorta, SMA = superior mesenteric artery.

The differences in the left and right sides of the DM are well understood at multiple levels of biology, but the maintenance of these asymmetries has not been explored. The classes of genes with asymmetric expression in the DM include transcription factors, ECM components, and genes which encode diffusible signals that could spread within the DM compartments to create morphogen gradients across the LR axis. Instead of a LR gradient, however, there is a sharp delineation between left and right DM in gene expression and protein localization, as well as LR asymmetric distribution of extracellular glycosaminoglycans, such as hyaluronan.33,36–40 Not surprisingly, perturbing even just one gene’s asymmetric expression patterns in the DM leads to aberrant gut looping patterns and abnormal vascular lesions.33,38–40 Thus, we hypothesized that a barrier exists in the center of the DM to segregate cells and diffusible signals to the left and right compartments. Here, we show that an atypical basement membrane forms a boundary between left and right cell populations and limits diffusion across the DM midline.

Results

Cells of left- and right-origin meet but do not mix in the DM

The DM mesenchyme forms by bilateral epithelial-to-mesenchymal transition (EMT) and ingression of coelomic epithelium, which is derived from the splanchnic mesoderm; the left DM comes from the left coelom and the right DM arises from the right coelom (Fig. 2A and B).35–37 This was visualized by injecting early chicken embryos with DiI and DiO into the right and left coelomic cavities, respectively (Fig. 2C-F), or by electroporating each side with plasmids encoding different fluorophores (Fig. 2G and H). While the left and right cells meet at the middle of the DM, they never cross over or mix.33,36–38,40,42–45 This striking separation occurs despite the lack of a visible histological boundary between the two sides as shown by H&E staining at HH20 and HH21, when the left is condensing and the right is expanding to drive the leftward gut tilting (Fig. 2K and L). Early in development the endoderm effectively separates the left and right splanchnic mesoderm (Fig. 2I and J), but once the DM forms and the endoderm descends it is likely important to continue separating the two sides until asymmetries can be established (Fig. 1B). In support of this, we have previously shown that when cell-cell adhesion is interrupted in the left DM, the cells become more dispersed36,38 and extend filopodia over towards the right side, suggesting pathogenic cell migration.38 Thus, the critical separation between left and right cells in the DM can be disrupted, necessitating a mechanism for protecting these asymmetries.

During DM formation from EMT and ingression of the coelomic epithelia, right and left cells do not mix.

When the coelomic cavity is injected with DiI at HH12-13, n = 5 (A), the labeled cells give rise to the mesenchymal and epithelial cells of the DM on the corresponding side of the embryo, n = 5 (B). When DiI and DiO are injected at HH12-13 into left and right coeloms, respectively, n = 6 (C), labeled cells are still segregated at HH18, n = 6 (D, E and F). The same results are found when cells are labeled by electroporation with pCAG-GFP (left) and pCl-H2B-RFP (right) (G and H), both when the midline is continuous (HH19 n = 3, G) and once it has disappeared (HH21 n = 3, H). (I-L) H&E staining of the DM at HH18 n = 5 (I) shows “empty space” between the notochord, endoderm, and dorsal aortae. At HH19 n = 5 (J), this space gains some cells (arrows), and the space is completely filled in by HH20 n = 4 (K), and HH21 n = 3 (L). Scale bars = 60 μm. nt = neural tube, c = coelom, ao = aorta, N = notochord, s = somite, DM = dorsal mesentery, L = left, R = right.

The DM midline is not marked by Lefty1, but by laminin

The early embryo uses a molecular barrier of Lefty1-expressing cells to separate laterality signals so the LR axis is established correctly.11–15 To see if this mechanism is adapted by the intestine for establishment of its laterality, we performed Lefty1 RNA in situ hybridization on both early (HH9) and later (HH19) stages. While Lefty1 was expressed at the midline of early embryos as expected (Fig. 3A), it was not expressed at the midline of the DM at HH19 (Fig. 3B). This indicates that a different mechanism must be at work during the establishment of gut asymmetries. Interestingly, scanning electron microscopy (SEM) data showed a fibrous matrix between the notochord and endoderm where the DM will later develop, suggesting that ECM may separate the two sides before they coalesce into the DM (Fig. 3C and D). Consistent with this hypothesis, basement membranes are found in other biological contexts where a barrier is needed, such as in the skin or around blood vessels.46 Basement membranes are dense ECM requiring laminin, collagen IV, nidogen, and perlecan and/or agrin (both heparan sulfate proteoglycans) with a large variety of other components that can be integrated to create specific “flavors” of basement membrane tuned to different barrier contexts.46 We therefore postulated that the DM midline has a physical barrier consisting of basement membrane, rather than a Lefty1 molecular barrier.

The DM midline is marked by laminin immunofluorescence.

(A) Lefty1 expression is seen at the embryonic midline of this HH9 embryo n = 4 (arrow). (B) Lefty1 expression is not seen at the midline of the DM (dashed box) of an HH19 embryo (n=4). (Notochord=positive control.) (C, D) Scanning electron microscopy (SEM) images of a fixed embryo at HH15-16 show fibrous ECM between the notochord and endoderm. C is from a more posterior axial level than D. (E-H) Time course of midline dynamics from HH18-HH21, marked by laminin. Scale bars = 25 µm. (I) Quantification of the intensity of laminin immunofluorescence over development, normalized to laminin staining around the neural tube. Biological replicates: HH18 n = 2, HH19 n = 3, HH20 n = 3, and HH21 n = 2. Each dot represents one image quantified. Statistical analyses are unpaired Welch’s t-tests. HH18-19: p = 0.0188, HH19-20: p = 0.0118, HH20-21: p = 0.0003. Additionally, there is a significant (p < 0.0001) linear trend among the means from HH18 to HH21 with a slope of -0.1019, r2 = 0.5906. (J) Midline appearance from hindgut to foregut in an HH19 embryo, marked by laminin. Scale bars = 25 μm. (K) Quantification of DM midline length (dashed line) of three HH19 embryos, from the separation of the notochord and endoderm (hindgut) to the fusion of the aortae (foregut, coinciding with midline fragmentation). (L) Model of DM midline time kinetics. N = notochord, E/en = endoderm, Ao = aorta, nt = neural tube, c = coelomic cavity, CE = coelomic epithelium, L = left, R = right.

To test this hypothesis, we visualized the basement membrane marker laminin by immunofluorescence (IF) for laminin alpha 1 (Lama1) at developmental stages where DM asymmetries are being established. As expected, this marked several typical, single-layer basement membranes underlying polarized cells, such as around the notochord (Fig. 3E and F),47,48 coelomic epithelium (Fig. 3E-H),49 and gut endoderm (Fig. 3E-H).50 We also observed scattered laminin staining in the DM mesenchyme, which is a consequence of those cells carrying basement membrane fragments with them after EMT and ingression from the coelomic epithelium.51,52 Interestingly, we identified a previously uncharacterized atypical double basement membrane within the DM. At HH18, when cellular DM asymmetries are first being initiated at the level of the midgut (which forms the small intestine), laminin IF marked an oval-shaped structure just ventral to the notochord and dorsal to the gut endoderm (Fig. 3E). No cells were seen within this structure as shown by a lack of nuclear staining (Fig. 3E) and empty space in SEM (Fig. 3C and DM, S1C). As the DM elongates and asymmetries become more apparent (HH19), this midline structure lengthened, forming two parallel lines connecting the endoderm and notochord (Fig. 3F). One stage later (HH20), the midline was still present but began to appear fragmented (Fig. 3G). By HH21, the asymmetries of the DM are established—the right DM has expanded, the left DM has condensed, the gut has tilted to the left, and vascular precursor cells have been driven out of the right side (Fig. 1B).33,40,43 Surprisingly, the midline disappeared by this stage, while laminin IF underlying the coelomic epithelium and endoderm remained intense (Fig. 3H). The lengthening of the midline and its subsequent loss occurred in an anterior-to-posterior wave down the embryonic gut tube (Fig. 3J and K). Consequently, anterior sections of younger embryos (i.e., HH12-13) had similar midline structures to posterior, older sections (i.e., HH18-19) (compare Fig. S2E-H with S2A-D).

The DM midline consists of a transient, true basement membrane

The combination of laminin, nidogen, perlecan (or agrin), and collagen is the foundation of all basement membranes.46 To further characterize the nature of the ECM at the DM midline, we did IF staining for nidogen and perlecan, confirming co-localization with laminin at the midline barrier (Fig. S3A, S3B). This further illustrates that the DM midline consists of basement membrane. This basement membrane structure is conserved in the squamate veiled chameleon, Chamaeleo calyptratus, which exhibits a similarly transient double basement membrane at the midline of the dorsal mesentery from approximately the 7-somite stage to the 29-somite stage (Fig. S5).53

In addition to the four foundational basement membrane components, a myriad of other proteins, proteoglycans, and glycoproteins can assemble onto the basement membrane.54 Consequently, there is a vast variety of “flavors” of basement membrane with different physical properties and different signals to adjacent cells about polarity, migration, or other behaviors.54 A common basement membrane constituent is fibronectin, which is best known for its role in the provisional matrix during wound healing.55 At our developmental stages of interest, fibronectin localized to the dorsal aorta which often coincides with the most ventral part of the midline (Fig. S2, S3C). Together, we model the midline as a transient double basement membrane that bisects the DM during developmental stages when critical asymmetries are being established (Fig. 3L).

The midline does not originate from the left or right DM

Although we have established the time kinetics of DM midline formation, the origin of this structure remains elusive. The midline is sandwiched between mesenchymal cells from the left and right DM, unbiased to either the left or the right side (Fig. 2F and G). Intriguingly, mesenchymal cells like these are not usually competent to construct an organized basement membrane.46,56 Mesenchymal cells can secrete matrix components,50 but the organization of these components into a basement membrane is dependent on the presence of cell surface anchors which are characteristic of tissues like polarized epithelium or endothelium, not mesenchymal cells.56 In the case of LAMA1, it is known that this protein is secreted by the epithelia in the developing intestine, not the mesenchyme.50 Indeed, RNA in situ hybridization for Lama1 did not show enriched expression in the mesenchyme at the DM midline (Fig. 4B and C). Moreover, if the cells adjacent to the midline were secreting and organizing the basement membrane, we would expect these cells to be polarized like the cells of the coelomic epithelium or endoderm. As expected, the left coelomic epithelium was polarized at HH19 relative to its basement membrane as quantified by Golgi staining with GM130 (Fig. 4D and E).38 However, GM130 staining and quantification showed that cells immediately to the left or right of the midline have random orientation (Fig. 4D and E). Together, these data allow us to rule out a mesenchymal origin for the DM midline.

The midline basement membrane is not made by the DM mesenchyme or EMT of the endoderm, and the notochord is not sufficient for midline formation.

(A-C) Lama1 RNA in situ hybridization and adjacent sections with laminin IF staining at HH16 n = 3, HH18 n = 9, and HH19 n = 10 (A’-C’). Scale bars = 50 μm. (D, E) Cell polarity analysis from GM130 staining shows that the mesenchymal cells immediately to the left or right of the midline (“left/right margin”) and within the double membrane (“midline cells” have random polarity, as do the cells of the right mesenchyme (random polarization control); in contrast to the strong apical-basal polarity in cells of the left coelomic epithelium. Five embryos were used for these quantifications. Number of cells per circle histogram: left coelomic epithelium = 209, left margin = 346, midline cells = 118, right margin = 413, right mesenchyme = 514, right coelomic epithelium = 295. (F) Electroporation mix containing pCAGEN-GFP plasmid was injected under an HH14/15 embryo and then electroporated to specifically target the endoderm. Lineage tracing endoderm-derived cells to HH19 n = 8 (G) and HH21 n = 2 (H) by pCAGEN-GFP electroporation of the endoderm. (I) Model of notochord transplant method. A piece of notochord (red) was isolated from an HH14 donor embryo. In a stage-matched recipient, a cut was made adjacent to the neural tube and the donor notochord was inserted into this slit. (J, L) RNA in situ hybridization for Chordin to mark the native notochord (blue dashed circle) and transplanted notochord (red dashed circle). (K, M) Laminin immunohistochemistry to mark basement membrane including the midline (white arrow). Notochords are marked with an N (native notochord) and N* (transplanted notochord). (J) and (K) are from the same embryo, as are (L) and (M). n = 8, scale bars = 50 μm. GT = gut tube, DM = dorsal mesentery, E = endoderm.

During endodermal descent, endodermal cells are not left behind to form the midline

Given that the DM midline connects the notochord and endoderm, both of which have their own basement membranes and are very closely associated early in development, we hypothesized that one or both of these structures contributes to midline production (Fig. 3C/E/J, S1A). As the embryo grows and the DM elongates, the distance between the notochord and endoderm increases (Fig. 2I-J, 3E-H), and the midline is found between them as a double line of basement membrane. Thus, we hypothesized that as the endoderm descends ventrally, it undergoes EMT and leaves behind basement membrane-carrying cells to form the midline. To test this, we developed a method to specifically target the endoderm using DNA electroporation (Fig. 4F). Briefly, we injected pCAG-GFP plasmid underneath HH14-15 embryos and applied an electric pulse such that the endodermal cells would take up the DNA (Fig. 4F), so we could lineage trace endodermal cells during DM formation. Interestingly, embryos isolated at HH19 and HH21 showed that GFP-labeled cells remain restricted to the endoderm—there were no GFP+ mesenchymal cells present in the DM (Fig. 4G, H). This indicates that the DM midline is not formed from EMT of basement membrane-carrying endodermal cells.

The notochord is not sufficient for DM midline formation

To test whether the notochord is sufficient for midline formation, we performed notochord transplant experiments. In brief, the notochord was removed from an HH12-15 embryo. An incision was made in a stage-matched recipient embryo adjacent to the neural tube and the donor notochord was inserted into this slit (Fig. 4I). Embryos continued to develop until isolation at HH19. These transplants were done to the left and right sides of different embryos (Fig. 4J/K and 4L/M respectively). RNA in situ hybridization was performed for chordin to ensure that the transplanted notochord was alive and functioning (Fig. 4J and L).57,58 Laminin IF did not reveal a secondary midline-like structure associated with the ectopic notochord (Fig. 4K and M) while the normal midline was unaffected (Fig. 4K and M, white arrows). This result is seen regardless of whether the transplants are done to the embryo’s left or right side. From this, we conclude that the notochord is not sufficient for the formation of DM midline.

The DM midline is resistant to degradation by Netrin4

Laminin matrices are susceptible to competitive disruption by the matrix protein Netrin4 (NTN4).59,60 NTN4 has very high binding affinity for laminin gamma subunits, such that NTN4 can prevent the formation of new laminin networks, which are the foundation upon which other basement membrane components assemble, and can also disrupt existing laminin networks.59,60 Ntn4 is not endogenously expressed in the DM (data not shown), which allows us to use it as a tool to target basement membranes in the DM. As expected, when we overexpressed Ntn4 on either side of the DM by electroporation, we perturbed the basement membrane underlying the coelomic epithelium and depleted the scattered laminin staining in the mesenchyme that results from EMT creating the DM (Fig. 5B, D, and E vs. controls 5A and C).51,52 Intriguingly, the DM midline basement membrane remained visibly intact in all embryos with lateral (Fig. 5B and D) and bilateral (Fig. 5E) Ntn4 overexpression in the DM. Similarly, endodermal Ntn4 overexpression caused much less disruption of the endodermal or midline basement membranes (Fig. 5G vs control 5F) when compared to its effect on coelomic epithelium basement membrane. This was true even when the electroporations were done much earlier in development (HH10 and HH12-13, data not shown). This suggests that the midline and endoderm may have basement membranes of the same “flavor,” possibly pointing to a common origin. In contrast, the basement membrane beneath the coelomic epithelium may be more susceptible to NTN4 disruption because of the prior EMT-induced breaks in the basement membrane,51,52 or a different protein composition.

Ectopic expression of Netrinn4 by electroporation visibly affects the basement membrane underlying the coelomic epithelium, but not that underlying the endoderm, or the midline.

(A, C) Electroporation of the left (A) or right (C) DM with the control, pCAG-GFP, had no effect on the basement membrane of the coelomic epithelium (arrows). Left n = 6. Right n = 4. (B, D) Electroporation of mouse Netrin4 (pCAGEN-Ntn4) and pCAG-GFP into the left (B) or right (D) DM disrupted the basement membrane underlying the coelomic epithelium (arrows) and scattered basement membrane in the mesenchyme (asterisks). Left n = 5. Right n = 5. The intact laminin staining in B is the basement membrane (BM) of the nephric duct. (E) Double DM electroporations also disrupt the coelomic epithelium (arrows) but the midline appears intact despite being contacted by Ntn4+ cells (n = 3). (F, G) Electroporation of the endoderm directly with pCAG-GFP or pCAGEN-Ntn4 does not visibly affect the basement membrane underlying the endoderm (open round pointers). Control n = 5. Ntn4 n = 8. The midline appears unaffected by any of these perturbations. Scale bars = 100 μm.

DM midline is a barrier against diffusion

Genes including Cxcl1233,39 and Bmp440 which encode diffusible signals are expressed asymmetrically in the DM (Fig. 6A and B). So too are genes encoding enzymes that are secreted into the ECM, like the HA-modifying enzyme TSG6.39 The expression domains of these genes have a sharp boundary at the midline, since left and right cells do not mix. However, the secreted protein products of these genes may be able to diffuse across the DM if their movement is not limited (Fig. 6C). We know that experimentally mixing left and right signals is detrimental to gut tilting and vascular patterning33,36–40—for example, ectopic expression of pro-angiogenic Cxcl12 on the right side results in an aberrant vessel forming on the right.33 Moreover, when the CXCR4 receptor antagonist AMD3100 (MW=502.78) is introduced to the left DM, it abolishes vascular development on the left. However, when the same drug is introduced to the right DM, the left-sided vascular development remains intact.33 This phenomenon suggests a barrier against diffusion.

The DM midline serves as a barrier against diffusion.

(A, B) Genes encoding diffusible signals including Cxcl12 and Bmp4 are expressed asymmetrically in the DM at HH19. (C) Hypothesis for the role of the midline in limiting diffusion of left and right signals across DM. (D) At HH19, the midline is intact (white arrow) and diffusion of 3000 MW dextran (green) is limited to the right side (n = 4/4). (E) At HH20, the midline (white arrow) has begun to fragment. Diffusion across the midline is prohibited in some embryos (n = 2/9) but permitted in others (n = 7/9). (F) At later stages when the midline has disappeared, diffusion is allowed through the DM (n = 7/7). (G) Proportion of dextran-labeled cells in the left vs. right DM, with unpaired t test. (H) At HH19, the midline is intact and diffusion of AMD3100-BODIPY is limited to the left side (n = 4). Dashed yellow box indicates quantified area. (I) At HH21 when the midline has disappeared, diffusion is allowed through the DM (n = 3). Dashed yellow box indicates quantified area. (J) Proportion of BODIPY intensity in the right vs. left DM, with unpaired t test. Each dot represents one image quantified. (K, L) Profile plot of average BODIPY intensity across the DM within the dashed yellow boxes in H and I, with left and right compartments of the DM overlayed. (M) Schematic of dextran injections into the right DM and AMD3100-BODIPY beading into the left DM. Scale bars = 50 um. LDM = left dorsal mesentery. RDM = right dorsal mesentery. GT = gut tube. E = endoderm. L = left. R = right. N = notochord. DA = dorsal aorta.

To test if the basement membrane structure at the midline is forming a functional barrier against diffusion, we injected 3 kDa fluorescent dextran directly into the right side of the DM (Fig. 6D-G). When these injections are performed at stages where the midline is intact (HH19), movement of dextran through the tissue was limited to the right side (n = 4/4 embryos) (Fig. 6D). When the basement membrane midline appears fragmented (HH20), these injections produced mixed results—in some embryos (n = 2/9), diffusion across the midline was prevented and in others (n = 7/9) the dextran was able to move into the left mesenchyme of the DM (Fig. 6E). Finally, at stages where no organized basement membrane structure remains at the midline (HH23), diffusion of dextran was always permitted across the entire width of the DM (n = 7/7) (Fig. 6F).

To confirm our finding that the basement membrane structure at the midline forms a barrier against diffusion, we utilized a BODIPY-tagged version of AMD3100,61 delivered via soaked resin beads surgically inserted into the left coelomic cavity (Fig. 6H-L). The ratio of average AMD3100-BODIPY intensity in the right DM versus the left DM was below 0.5 when the midline is intact (HH19, n = 4/4), indicating little diffusion across the DM (Fig. 6J). When no midline remains at a later stage (HH21, n = 4/4), this ratio significantly rises to near one, indicating diffusion of the drug is not impeded when the midline basement membrane structure is absent (Fig. 6J). Collectively, these data suggest that the basement membrane structure at the midline forms a transient functional barrier against diffusion (Fig. 6M).

Discussion

Establishing the left and right body plan in the early embryo is a fundamental part of development and this process depends on the presence of a Lefty1+ midline barrier. Individual organs, too, have LR asymmetries, but the brain has the only known organ-specific midline barrier, where commissural axons of the brain and spinal cord are tightly controlled by midline-localized guidance and repulsion cues including FGFs, SLIT/ROBO signaling, EFNB3, heparan sulfate proteoglycans, and the Rac-specific GTPase-activating protein α-chimaerin.22–28 The developing intestine has a similar need for separation between left and right cells and signals, but it seems to accomplish this by a different mechanism—an atypical basement membrane at the midline.

This basement membrane may separate cells that have ingressed from the right and left coelomic epithelia,62 since these cells do not mix at the midline (Fig. 2). It may also prevent mixing of diffusible signals. While the diffusion of a given signal depends on the tissue context,63 some morphogens (like Nodal) can induce effects at distances of 200 μm or greater.64 The HH19 DM is only about 150 μm across, not a prohibitive distance for diffusible signals like BMPs, TGFß,40 and CXCL1233 to cross between the left and right DM. Thus, we hypothesized that the DM may need a barrier at the midline to segregate these signals. In support of this, we showed that the midline limits diffusion of dextran from right to left, which suggests that it also blocks the movement of endogenous diffusible signals. Moreover, with a molecular weight of just 3 kDa, dextran’s inability to cross the midline indicates that diffusible proteins of typical weight (such as CXCL12 at 10 kDa and BMP4 at 34 kDa) are also unable to cross. We previously showed evidence that even the 502.78 Da drug AMD3100 was prevented from moving to the other side of the DM.33 In line with these findings, we have now employed a BODIPY-labeled form of AMD310061. Similar to dextran, our results show BODIPY diffusion across the midline when the basement membrane is intact. However, when no basement membrane remains at the midline, BODIPY diffusion was permitted across the entire DM width. While the midline could not be degraded by Ntn4 overexpression, future studies may reveal tools for the selective destruction of this basement membrane to better understand its function in gut laterality.

This work adds a new facet to our knowledge of basement membrane form and function. Basement membranes play many critical barrier functions in the embryo and adult, usually found as a single layer that underlies polarized epithelial or endothelial cell layers such as those lining the intestines, encircling blood vessels, or enveloping muscle cells, adipocytes, or Schwann cells.46 As such, null mutations in genes encoding basement membrane components often result in embryonic lethality and postnatal pathologies.65–70 However, a role for basement membrane in establishing LR asymmetry has not been described previously.

The basement membrane we describe here is atypical in its double membrane structure, which raises interesting questions about its formation. We showed that the midline is not produced by the mesenchymal cells of the DM (Fig. 4), and that the notochord is not sufficient for its synthesis (Fig. 4). Instead, we consider the endoderm. Upon electroporation of the endoderm with GFP, GFP-positive cells were not detected in the DM later in development but remained restricted to the endoderm (Fig. 4F-H), indicating that the midline does not form from EMT of basement membrane-carrying endodermal cells. We can also eliminate endodermal death as a possible mechanism, because there was no appreciable cell death observed by TUNEL staining at the midline during the stages of interest (data not shown). Nor does the midline form in the same way as other double basement membranes, which arise from the meeting of the basal sides of two tissues.71,72 This is classically illustrated in the kidney glomerulus between epithelial podocytes and endothelial cells72–74 and in the blood-brain barrier between endothelial cells, pericytes, and astrocytes.71,75 Instead in the case of the DM midline barrier, the apical sides of the endodermal cells are facing each other.

We hypothesize that the midline forms when basement membrane is left behind as the endoderm descends ventrally during normal development, as if it were a “scar” of where the endoderm was previously (Fig. 7). During early development the notochord is embedded within the endoderm and a basement membrane covers the two structures. Only later are the two structures separated by a full basement membrane, pointing to a strong connection between the two.76,77 The medial migration of the aortae and coelomic epithelia could provide tissue forces that push the notochord and endoderm apart, leaving behind basement membrane at the DM midline. The observation that the basement membranes of both the endoderm and midline were resistant to disruption by NTN4 could also support the idea that the endoderm is responsible for making both basement membranes (Fig. 7). In support of this, we know that the basement membrane underlying the gut endoderm does not co-migrate with the intestinal epithelial cells as they move from the proliferative intestinal crypts to the tip of the villus over the course of three to six days in the adult; the basement membrane is instead left in place.78 One way to test this hypothesis would be to “trace” endodermal basement membrane by electroporation of a tagged basement membrane component, as has been done recently in C. elegans79–84 and Drosophila.85,86 However, these advances are only just now reaching the mouse system87,88 and have not yet made their way into the chick embryo.

Model of endoderm descending hypothesis for midline formation.

We hypothesize that as the endoderm moves ventrally and the distance between the notochord and endoderm grows, basement membrane from the endoderm may be left behind. This can be compared to a zipper where each side is the basement membrane underlying the endoderm, and when the zipper pull (tip of endoderm) moves downwards, the basement membrane behind it pulls closer together.

The endodermal basement membrane and midline’s resistance to NTN4 disruption is particularly notable, as similar resistance has not been reported before. This suggests the presence of an unknown factor or modification that shields or stabilizes the laminin network from NTN4 binding. Investigating the mechanisms that stabilize basement membranes against such disruptions could be important for future developmental studies and research on basement membranes in the context of disease and cancer.

Although Lefty1 expression is not observed at the midline when this basement membrane is robust, it remains unclear whether midline Lefty1 expression is crucial for the construction or proper function of this structure. Future research could involve siRNA inhibition of Lefty1 expression at an early stage, followed by an evaluation of the midline structure’s integrity at HH19. However, similar to perturbing global Nodal expression, it may be challenging to differentiate between the direct impact of Lefty1 loss on future midline cells, and the indirect impact on midline cells resulting from free Nodal diffusion throughout the early lateral mesoderm that would happen as a consequence of Lefty deficiency. Further investigation into the relationship between these two aspects of the midline barrier is important.

The midline barrier is set apart from many other basement membranes by its rapid disappearance. We consider that the loss of the DM midline barrier may be caused in part by “stretching” of the basement membrane as the notochord and endoderm become increasingly separated due to the elongation of the DM (Fig. 3J). Since the DM itself does not contribute basement membrane to the midline (Fig. 4C) and the midline length increases quickly (Fig. 3J), it is plausible that the midline may be pulled until it reaches its ultimate tensile strength (between 0.5-3.8 MPa in other examples of naturally occurring basement membranes),89 and then breaks. This would suggest that midline breakdown is a passive consequence of embryo growth.

However, we also consider that there could be an active breakdown mechanism for the midline. The turnover of stable basement membranes occurs on the scale of weeks,78,90 but the midline barrier appears to degrade over just 12-24 hours. In other contexts, basement membrane destruction can occur over a large area, such as by secretion of matrix metalloproteinases (MMPs) in metastasizing cancer,91,92 or more localized, as exhibited by invadopodia on immune or cancer cells before metastasis.93,94 Such localized basement membrane breakdown is seen in critical developmental processes, including mouth development in deuterostome embryos.95 The oral membrane includes a basement membrane that closes off the digestive system from the outside world. This basement membrane specifically disintegrates to rupture that membrane to open the early mouth cavity.95 In addition, a basement membrane divides the two halves of the embryonic brain, and must be broken down at the site of the corpus callosum for neurons to cross for inter-hemisphere communication in the cerebrum.96,97 Localized basement membrane dissolution is also critical for optic cup fusion, where a contact-dependent dissolution of basement membrane occurs between the two sides of the optic fissure.98,99 Coloboma, a congenital eye defect where some tissue is missing inside the eye resulting in an enlarged, irregular pupil, occurs when optic fusion arrests.100,101 It is worth noting that in all of these examples and the case of the midline, the location of basement membrane breakdown is very specific—neighboring basement membranes appear unaffected (Fig. 3H). Together, these findings support the idea that basement membrane breakdown in the DM may not simply be a passive process but may be required by the embryo for later events in gut development or vascular patterning. We have not yet identified any matrix protease specific to the midline of the DM. Given that a mechanism for basement membrane breakdown is not well characterized in any of the above contexts, there is high potential for future investigation.

The DM midline may have effects on development beyond the intestine. It has long been known that signals produced from axial structures in the embryo are critical for the establishment of LR asymmetry, but it has been difficult for researchers to pin down which “midline structure” or “midline signal” is actually responsible. For example, the heart,102,103 lungs,44 and kidneys104 all rely on “dorsal midline structures” to develop properly. Aberrations from the loss of midline structures like the notochord include “horseshoe kidney,” in which the kidneys stay close to the midline and fuse at their posterior end.105 This is due to the lack of notochordal Shh signaling, but the “midline barrier” that exists downstream from this Shh signaling remains a mystery.106

The DM midline barrier may also play a role in vascular patterning, including aortic fusion. The aorta begins as two parallel tubes with an avascular zone in between (where the midline is),107 but progressively they fuse into one with an anterior-to-posterior wave until the level of the vitelline arteries (Fig. S2A-D).107 Fusion of the aortae coincides with the fragmentation and disappearance of the midline, suggesting that these two processes may be inextricably linked (Fig. 3J, S2). Proper timing of this fusion depends on carefully balanced levels of VEGF,108 SHH,109 and the anterior-to-posterior wave of downregulation of the BMP-inhibitory genes Chordin and Noggin from the notochord.107,110,111 The precise mechanism of dorsal aorta fusion remains unknown, although there is evidence that VEGF signaling pulls VE-cadherin away from its cell-cell junctions.108 This relocalization may be important for the remodeling of the aortic endothelium during fusion and may be linked to the breaking of the DM midline barrier.

Midline structures are critical for laterality to develop correctly. The notochord is certainly involved, particularly as the source of modulators of BMP and Hedgehog signaling. However, this does not seem to be the whole story, given that in each of these contexts the actual “midline barrier” downstream from notochord signals has not been identified. It is possible that the DM midline basement membrane is key here, either for separating left and right signals or perhaps for binding signals69 from the notochord to act as a buffer between each side. The midline barrier may also play a role in the rheology of the DM. Microindentation analyses at HH21 show that the condensed left DM is significantly stiffer than the expanded right DM, and that proper gut tilting is dependent on this difference being tightly regulated.40 It is possible that the midline barrier helps to segregate the stiffness-influencing components of each side (i.e. covalently modified HA on the right,39 N-cadherin on the left36) and provides a “wall” for the right side to push upon in order to swing the gut tube towards the left. It’s important to note that left and right mesenchymal cells do not cross the midline to the opposite side of the DM. Likewise, during the formation of gut arteries, a small subset of gut vascular endothelial cells migrate from right to left, closely adjacent to the dorsal tip of the endoderm, but they do not traverse the midline.33 Moreover, when the midline basement membrane is present and intact, N-cadherin expression is symmetric across the DM, but becomes asymmetric and left-specific upon midline disintegration.36 This may represent a shift in cell segregation mechanisms in the DM - while the midline is intact, the double basement membrane structure appears sufficient to segregate left and right cells. When the midline disappears, a new cellular separation mechanism is required to maintain asymmetric compartments, and N-cadherin functions in this role.

Looping patterns of the midgut are stereotypical between individuals but vary between species.31 While differential growth rates between the gut tube and the dorsal mesentery have been identified as the primary means by which distinct gut looping patterns may be achieved,31 it is also possible that mesentery-specific changes such as modifications to the kinetics of asymmetry in the DM may impact looping patterns with potentially adaptive impact on diet and therefore niche utilization. The presence, permeability, and timing of degradation of the midline basement membrane may provide evolution with an additional means of fine-tuning looping patterns between species. We have found this basement membrane to be transiently present in the DM of veiled chameleon embryos from just after the seven-somite stage to the 29-somite stage (Fig. S5),53 suggesting this structure is conserved in reptiles and birds at a minimum, with intriguing heterochrony in degradation between species. In mammals, prior studies have observed midline laminin deposition in the mouse gut located between the separating notochord and endoderm, further suggesting conservation of a midline basement membrane in amniotes.112,113

Collectively, we have identified a novel midline barrier in the gut mesentery that is composed of an atypical double basement membrane that forms a boundary between the left and right sides and limits movement of diffusible signals and cells, at a stage when Lefty1 is no longer expressed at the midline. The DM midline also presents an opportunity to interrogate the fundamental mechanisms of basement membrane formation and degradation during vertebrate embryonic development, with implications for research on cancer metastasis. We posit that this midline is a distinct strategy for the critical separation of left and right signals and cells, key for establishing and maintaining LR asymmetry for healthy gut development.

Materials and methods

Chicken embryo development and processing

Fertile chicken eggs were purchased from Westwind Farms (Interlaken, NY http://chickenhawkfood.com). After 36-48 hours of incubation at 37°C, eggs were windowed by removing 8 mL of thin albumen with an 18 ½ gauge needle/10 mL syringe and cutting an oval in the side of the shell, then covering the opening with clear packing tape and returning the egg to the incubator. Once at the desired stage, embryos were isolated in cold 1X PBS and fixed overnight in 2% paraformaldehyde (PFA) at 4°C, followed by PBS washes. Embryos were prepared for cryo-embedding by putting them through graded sucrose solutions ending in 30% sucrose overnight at 4°C. Embryos were cryo-embedded in OCT (VWR 25608-930), sectioned to 15 μm, dried overnight, then stored at -80°C.

Immunofluorescence (IF)

Cryosections were rehydrated in PBS then PBST (0.03% Tween-20), then blocked in 3% heat-inactivated goat serum (HIGS, Gibco 16210072) in PBST for 45 min at room temperature. Primary antibodies were diluted in blocking solution (3% HIGS in PBST) and incubated either for 45 minutes at room temperature or overnight at 4°C. After three PBST washes for five minutes each, secondary antibodies were incubated for 45 minutes at room temperature with 1:2000 dilution of DAPI added. PBST and PBS washes were done before mounting the slides with Prolong Gold anti-fade (Invitrogen P36930). Antigen retrieval pretreatment was necessary for GM130 (BD Biosciences 610822) immunofluorescent staining. Cryosections were rehydrated in water, then microwaved in 1:100 antigen-retrieval solution (Vector Laboratories, H-3300) until nearly boiling. After incubating at 37°C wrapped in aluminum foil for 15 minutes and cooling for 10 minutes, slides were taken through the standard IHC protocol.

RNA in situ hybridization

Section and wholemount RNA in situ hybridization was done using a modified protocol from Moisés Mallo as previously described.114

Dextran injections

Dextran injections into the DM were done using 3000 MW dextran conjugated to fluorescein (ThermoFisher D3306) at a concentration of 10 mg/mL in 1X PBS with Fast Green dye added to better visualize the solution during injection. This mixture was loaded into fine pulled glass capillary needles. A microinjector with a foot pedal was set to 5 psi for 200 ms. Chicken embryos at the desired stage were prepared by removing the vitelline membrane. Injections were done to the right DM only, because embryos lie on their left sides from HH18 onwards and only the right DM is accessible for injection. With the anterior/posterior axis of the embryo perpendicular to the needle and with the needle at a 25° angle, the body wall was gently pulled back so the needle could access the right side of the DM. The needle was gently pressed into the tissue until the embryo moved slightly from the force. Then the foot pedal was pressed once to inject. Embryos were allowed to continue incubating for about two hours, then embryos were collected and fixed in 2% PFA overnight at 4°C. To screen for embryos with quality injections, embryos were cryo-embedded and sectioned. Any embryos with visible damage to the DM in these sections were excluded from further analysis.

AMD3100-BODIPY beading

AMD3100-BODIPY was synthesized as previously described,61 and AG beads (Bio-Rad 143-1255) were soaked in 5 mg/ml AMD3100-BODIPY overnight nutating at 4°C. Drug-laden beads were inserted into the left coelomic cavity via a small incision, as previously described.33

CM-DiI and SP-DiO injections

Five μl of a stock solution of CM-DiI or SP-DiO (1μg/μl in EtOH) was diluted into 45 μl of prewarmed 0.3M sucrose in single distilled water maintained at 37°C. Dye solutions were injected into the coelomic epithelium as previously described.44

Electroporation

DM electroporations were performed as described previously.39,115 Endodermal electroporations were performed with a similar method, but with the electroporation mix (plasmid of interest and/or pCAG-GFP, 1X PBS, 1X Fast Green, 1mM MgCl2, and 0.17% carboxymethylcellulose) injected into the empty space beneath the ventral side of the embryo while the negative electrode was held in place there. The positive electrode was placed directly above the negative electrode, centered along the neural tube, before the pulse was applied. pCAGEN-Ntn4 expression plasmid was constructed by cloning the full-length coding sequence out of mouse Ntn4-AP-His plasmid (Addgene 71980) using the primers in Table 1, and cloning into pCAGEN (Addgene 11160) with XhoI (NEB R0146) and NotI (NEB R0189). A second Ntn4 construct was also used, using a mouse Ntn4 (generously provided by Raphael Reuten) and cloned into the pMES vector,116 with similar results.

Notochord transplants

Notochord transplants were performed on HH12-HH15 embryos, using a method adapted from papers describing notochordectomies and notochord transplants.117–123 Note that notochordectomies, while potentially informative, were not done because damage to the endoderm is highly likely in those experiments. Since the endoderm is also potentially implicated in midline formation, an experiment that did not perturb that tissue was preferred. To prepare the donor notochord, the embryo was cut crosswise at the level of the vitelline arteries and close to the end of the tail. Clean 2-4 mm sections of notochord were used for the transplants. A sharp glass needle was used to make an incision in the recipient embryo along the anterior/posterior body axis between the neural tube and somites, deep enough so the ectopic notochord could sit next to the native notochord without puncturing the dorsal aorta. The donor notochord was pressed into place using a pair of dull glass needles or forceps. Embryos continued developing at 37°C until stage HH19.

Scanning electron microscopy (SEM)

Embryos used for Fig. 3C and D were fixed in 4% paraformaldehyde and 2% glutaraldehyde in PBS overnight at 4°C. Samples were then equilibrated in 5% sucrose/PBS for 1 hour at RT, 20% sucrose/PBS for 1 hour at RT, and finally 15% sucrose/7.5% gelatin/PBS at 37 degrees overnight. Embryos were then embedded in plastic molds and frozen in liquid Isopentane in a dry ice-ethanol bath. Cryosections of 10 microns in thickness were collected on poly-L-lysine treated coverslips and incubated twice with fresh drops of PBS for 10 min at 37°C. Coverslips were then washed with 0.1M cacodylate buffer and post fixed with 0.1% osmium tetroxide. Following washes with deuterium-depleted water (DDW), the sections were dehydrated in graded ethanol series, critical point dried (Quorum K850) and sputter coated with 6 nm of chromium (Quorum Q150T). Samples then were viewed on Zeiss Ultra Plus HR Scanning Electron Microscope using the SE2 detector.

Embryos for SEM in Fig. S1 were fixed in 2% glutaraldehyde in 0.50M cacodylate buffer (pH 7.4) at 4°C for two hours, then rinsed three times for 10 minutes each in 0.05M cacodylate buffer. At this point, embryos were cut down to size, using sharp spring scissors to cut the embryo crosswise at the level of the midgut. Then embryos were post-fixed in 1% osmium tetroxide in 0.05M cacodylate buffer at 4°C for one hour, rinsed again in 0.05M cacodylate buffer (3x 10 min), dehydrated in an ethanol series of 25%, 50%, 70%, 95%, and 100% for 10 minutes each, and left in 100% ethanol overnight. The following day embryos were critical point dried in CO2, soaking for 24 hours. Samples were mounted and silver paint was used for conductivity. These embryos were sputter-coated with gold palladium and imaged on a LEO 1550 (Keck SEM).

Imaging, image processing, and quantifications

Brightfield and fluorescent images of tissue sections were taken on a Zeiss Observer Z1 with Apotome, a LSM880 Confocal multiphoton inverted microscope—i880 (Zeiss), or a LSM710 Confocal (Zeiss), or a ScanScope CS2. Stereoscopic images were taken on a SteREO Discovery.V12 (Zeiss). Images were processed using Fiji. Statistical analyses were done using GraphPad Prism.

Quantification of fluorescent intensity of laminin staining was done for five sum intensity projections per biological replicate. For each image, the width of the midline was averaged from three measurements. When no midline is discernible at HH21, the midline width for each image was replaced with the overall average HH20 midline width. Then, five profile plots of gray values (intensity) were obtained from orthogonal lines drawn across the midline. The gray values within the average membrane width centered around the local maximum gray value were averaged to produce a raw mean midline gray value for each image. This process was repeated for neural floor plate basement membrane, and the raw mean midline gray value for each image was normalized to the raw mean floor plate basement membrane value. The means of HH18, HH19, HH20, and HH21 normalized midline gray values were compared via Welch’s unequal variances t-test. Finally, midline length was measured at HH19 as the distance between the tip of the endoderm and the notochord from the hindgut (where the “bird’s nest” of laminin deposition becomes longer than it is wide), to the cranial midgut (where the branches of the dorsal aorta fuse).

AMD3100-BODIPY diffusion was quantified by drawing an ROI across the DM bound by the dorsal aorta on top and the endoderm on the bottom. A profile plot of average gray values was obtained for each ROI for laminin and AMD3100-BODIPY separately. At HH19, the x-positions of laminin maxima corresponding to the left and right coelomic epithelium basement membrane and the midline basement membrane were used to divide the BODIPY intensity data into left and right DM compartments. At HH21 when no midline is present, left and right DM compartments were designated by dividing the distance between coelomic epithelial basement membranes in half. The separated left and right BODIPY data were then averaged to obtain a mean left DM and mean right DM BODIPY intensity value, which were compared to obtain the ratio of right versus left DM BODIPY signal. All intensity measurements were normalized for each image by subtracting the average gray value of the notochord in the 488 channel. These ratios were compared via Welch’s unequal variances t-test.

GM130 Golgi staining was used to assess cell polarity as described previously.38 Five cell populations were evaluated: cells contacting the midline on the left (“left margin”), cells contacting the midline on the right (“right margin”), right mesenchymal cells not contacting the midline or coelomic epithelium, left coelomic epithelial cells, and rare cells observed between the double membrane of the midline itself (“midline cells”) (Fig. 4D, Fig. S4). For each cell, the clockwise angle relative to vertical (0°) of the line drawn between the center of the nucleus and the Golgi apparatus (Fig. 4D inset) was recorded and plotted on an angle histogram in 20° bins with five biological replicates each. The trend of cell polarity in each cell population was assessed using Rayleigh’s test of uniformity using the “circular” R package for circular statistics (https://r-forge.r-project.org/projects/circular/).

Veiled chameleon husbandry, collection, and fixation

Veiled chameleon husbandry was performed at the Stowers Institute for Medical Research in the Reptiles and Aquatics Facility in accordance with the Institutional Animal Care and Use Committee approved protocol 2020-115, and as described previously.124–126 following the protocols which are publicly available here: dx.doi.org/10.17504/protocols.io.bzhsp36e. Veiled chameleon eggs were collected at oviposition in the Reptiles and Aquatics Facility at Stowers Institute for Medical Research. Eggs were incubated in deli cups with moist vermiculite at a constant temperature of 28 °C for 65-90 days to achieve desired staging. The eggs were cleared of large particles and wiped with RNaseZap wipes (Invitrogen AM9786) to minimize RNAse contamination. Clean eggs were candled to determine the position of the embryo under the leathery shell. We used fine scissors to cut a segment of the shell around the embryo and separate the embryo (attached to the shell) from the rest of the egg. The embryos were further separated from the shell and dissected out of the membranes in room temperature Tyrode’s solution, made in DEPC-treated water. Subsequently the embryos were fixed overnight at 4 °C in 4% PFA in DEPC 1X PBS, then dehydrated through an ascending methanol series into 100% methanol and stored at -20 °C for future analysis. Stages of embryonic development were determined as previously described.53